---
title: "002 — Baseline classifiers (mlr3) on the labelled SNI feature matrix"
author: "Dan MacLean"
date: "2026-04-23"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  message: false
  warning: false
---

## Goal

Quantify how well the canonical SNI features distinguish helper from
sensor proteins on the 18-protein labelled benchmark, using an mlr3
pipeline under leave-one-out cross-validation (LOOCV). Three feature
subsets — Scientist, Co-Scientist, and Combined — are evaluated under
two strategies for the α1-NA features (drop them, or median-impute
them). Observed LOOCV accuracies and AUCs are then compared against
permutation null distributions obtained by re-running the same
pipeline on label-shuffled data.

The idea here is that because we're setting up an `mlr3` factory-style
classifier pipeline, we can be maximalist about what sub-sets/super-sets
to look at. We can therefore compare the Co and Scientist features at low effort
and the effect of the missing value features.

We are optimistic for good classifiers because of the prevalance of single
features that do excellent sensor/helper classification on there own.

## Hypotheses

1. Multiple single features already give perfect rank separation on
   this benchmark (notebook 001), so multivariate LOOCV accuracy will
   be at or near 1.0 for at least the Scientist and Combined subsets
   under either α1-policy.
2. The Scientist and Co-Scientist subsets, evaluated on the same 18
   proteins, will give LOOCV scores that are not distinguishable from
   each other, since one misclassification at n = 18 corresponds to a
   change of 1/18 ≈ 0.056 in accuracy.
3. The permutation null distribution of LOOCV accuracy will be roughly
   centred on 0.5; empirical p-values for the observed scores will be
   bounded by 1 / (n_perms + 1) when no shuffled labelling matches.

## Rationale

n = 18 is small. We use (a) LOOCV so every sample serves once as
test, (b) more than one learner so the result is not a property of a
single model class, (c) explicit α1-feature handling — drop vs
median-impute — to separate the contribution of the structurally
missing features from that of the always-observed features, and (d)
a permutation null so observed scores can be read against the
distribution produced by the same pipeline on randomly shuffled
labels.

------------------------------------------------------------------------

## Setup

```{r setup}
suppressPackageStartupMessages({
  library(mlr3)
  library(mlr3learners)
  library(mlr3pipelines)
  library(here)
  library(ggplot2)
  library(tidyr)
  library(data.table)
})
lgr::get_logger("mlr3")$set_threshold("warn")

knitr::opts_chunk$set(fig.path = "../results/figures/002_")

mat_11 <- readRDS(here("data", "feature_matrix_labelled_11.rds"))
mat_11$class <- factor(mat_11$class, levels = c("sensor", "helper"))

dir.create(here("results"), showWarnings = FALSE, recursive = TRUE)
```

## 1. Feature subsets and design cells

The "Scientist", "Co-Scientist" and "Both" subsets are taken from
`SolNRCH_foldome/R/NRC_heatmap_analysis.R`. The α1-NA set is the three
columns shown in notebook 001 to be NA in 9/9 sensors.

```{r design}
sci_full   <- c("ipTM_LCB", "Sum_CONTACTS", "S_PROTO", "L_APEX_mean")
cos_full   <- c("D_MHD_P_mean", "THETA_APEX_mean", "H_ABS_mean", "MU_H_mean")
both       <- c("BSA_INT_PROTO", "SD_THETA_ROT", "D_APEX")
combined   <- unique(c(sci_full, cos_full, both))     # 11 features

alpha1_na  <- c("THETA_APEX_mean", "L_APEX_mean", "MU_H_mean")

make_cell <- function(name, features, impute) {
  list(name = name, features = features, impute = impute)
}

design <- list(
  make_cell("scientist_drop_a1",     setdiff(sci_full, alpha1_na),   FALSE),
  make_cell("scientist_impute_a1",   sci_full,                       TRUE),
  make_cell("coscientist_drop_a1",   setdiff(cos_full, alpha1_na),   FALSE),
  make_cell("coscientist_impute_a1", cos_full,                       TRUE),
  make_cell("combined_drop_a1",      setdiff(combined, alpha1_na),   FALSE),
  make_cell("combined_impute_a1",    combined,                       TRUE)
)

design_summary <- do.call(rbind, lapply(design, function(d) {
  data.frame(
    cell        = d$name,
    impute_a1   = d$impute,
    n_features  = length(d$features),
    features    = paste(d$features, collapse = ", ")
  )
}))
knitr::kable(design_summary)
```

## 2. Pipeline helpers

A small set of helpers wraps the mlr3 task / learner / resample
boilerplate so each design cell × learner combination is one function
call.

```{r helpers}
make_task <- function(data, features, target = "class") {
  d <- as.data.frame(data[, c(target, features), drop = FALSE])
  TaskClassif$new(id = "sni", backend = d, target = target,
                  positive = "helper")
}

build_learner <- function(lrn_id, impute) {
  base <- lrn(lrn_id, predict_type = "prob")
  if (lrn_id == "classif.ranger") base$param_set$values$num.trees <- 200
  if (impute) {
    g <- po("imputemedian",
            affect_columns = selector_type(c("numeric", "integer"))) %>>% base
    GraphLearner$new(g)
  } else {
    base
  }
}

run_loocv <- function(data, features, lrn_id, impute) {
  tsk <- make_task(data, features)
  lrn_obj <- build_learner(lrn_id, impute)
  rr <- suppressWarnings(resample(tsk, lrn_obj, rsmp("loo")))
  # AUC is undefined per LOO fold (one observation each); score on the
  # pooled prediction across folds so the 9-helper / 9-sensor ranking is
  # used as a single sample.
  acc_mean <- rr$aggregate(msr("classif.acc"))
  auc_pool <- rr$prediction()$score(msr("classif.auc"))
  list(rr = rr,
       agg = c(classif.acc = unname(acc_mean),
               classif.auc = unname(auc_pool)))
}
```

## 3. Baseline benchmark

Three learners types — `featureless` (predicts the majority class),
`log_reg` (plain logistic regression), and `ranger` (random forest,
200 trees) — are run on each of the six design cells. `featureless`
gives the no-information baseline by bascically calling just the class with the most members as its prediction for everything `log_reg` and `ranger` give two
contrasting model classes (linear and non-linear ensemble).

```{r benchmark}
LEARNERS <- c("classif.featureless", "classif.log_reg", "classif.ranger")

set.seed(1)
bench_results <- do.call(rbind, lapply(design, function(cell) {
  do.call(rbind, lapply(LEARNERS, function(lrn_id) {
    res <- tryCatch(
      run_loocv(mat_11, cell$features, lrn_id, cell$impute),
      error = function(e) list(agg = c(classif.acc = NA_real_,
                                       classif.auc = NA_real_))
    )
    data.frame(
      cell    = cell$name,
      impute  = cell$impute,
      learner = sub("^classif\\.", "", lrn_id),
      acc     = unname(res$agg["classif.acc"]),
      auc     = unname(res$agg["classif.auc"])
    )
  }))
}))

bench_results
```
Ok, so our benchmark is helpful. `featureless` scores 0 — not a chance result but an LOOCV artefact: with balanced classes, holding out a helper flips the training majority to sensor (and vice versa), so featureless is wrong every fold. Read it as "no feature information used" rather than a literal 0% accuracy. Past that, the headline is `combined` + `ranger`: both the drop and impute cells hit perfect LOOCV accuracy and AUC, so the full 8–11 feature set is where the real classifier lives. The Scientist-only and Co-Scientist-only subsets are both strong but trail combined by a misclassification or two. The models discern sensors and helpers pretty well.


```{r benchmark-table}
wide_acc <- pivot_wider(bench_results[, c("cell", "learner", "acc")],
                        names_from = learner, values_from = acc)
wide_auc <- pivot_wider(bench_results[, c("cell", "learner", "auc")],
                        names_from = learner, values_from = auc)
knitr::kable(wide_acc, digits = 3, caption = "LOOCV accuracy")
knitr::kable(wide_auc, digits = 3, caption = "LOOCV AUC")
```
Same thing in a wider table.

## 4. Held-out predictions per protein

For the two "combined" cells (drop α1 vs impute α1), report the
LOOCV-held-out predicted probability of `helper` for each protein,
under the random-forest learner. This shows which proteins, if any,
are misclassified or sit close to the 0.5 boundary. Basically, we'll leave out each protein in turn,
retrain and see what its $p(helper)$ is.

```{r held-out}
held_out <- function(data, features, lrn_id, impute) {
  tsk <- make_task(data, features)
  lrn_obj <- build_learner(lrn_id, impute)
  rr <- suppressWarnings(resample(tsk, lrn_obj, rsmp("loo")))
  pred_dt <- as.data.table(rr$prediction())
  pred_dt$ID    <- data$ID[pred_dt$row_ids]
  pred_dt$class <- data$class[pred_dt$row_ids]
  pred_dt[, .(ID, class, response, prob_helper = prob.helper)]
}

ho_drop   <- held_out(mat_11, setdiff(combined, alpha1_na),
                      "classif.ranger", impute = FALSE)
ho_impute <- held_out(mat_11, combined, "classif.ranger", impute = TRUE)

ho_drop$cell   <- "combined_drop_a1"
ho_impute$cell <- "combined_impute_a1"
ho_all <- rbind(ho_drop, ho_impute)
ho_all <- ho_all[order(ho_all$cell, ho_all$class, -ho_all$prob_helper)]
knitr::kable(ho_all, digits = 3)
```

```{r held-out-plot, fig.width=8, fig.height=5}
ggplot(ho_all, aes(x = prob_helper, y = reorder(ID, prob_helper),
                   colour = class, shape = response == class)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.5) +
  facet_wrap(~ cell) +
  scale_colour_manual(values = c(sensor = "#e76f51", helper = "#264653")) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 4),
                     labels = c(`TRUE` = "correct", `FALSE` = "wrong"),
                     name = "prediction") +
  labs(x = "LOOCV held-out P(helper) (ranger)", y = NULL) +
  theme_minimal(base_size = 10)
```
They're all re-predicted correctly and most $p(helper)$ stays appreciably away from the 0.5
zone of not being so sure.

## 5. Permutation null distribution

Class labels are permuted across the 18 proteins and the LOOCV
pipeline is re-run on each shuffle. `n_perms = 200` shuffles per
cell, ranger learner only (chosen for speed; the same approach
applies to the other learners). Empirical one-sided p-values are
`(1 + sum(null >= obs)) / (n_perms + 1)`.

Standard 'bootstrap-like' shuffling of labels for 200 reps.
The idea being that we should see a nice aggregation of
LOOCV accuracy (and AUC) around 0.5. Which would indicate
that the random shuffling makes the model unable to recall
the original labels. Randomly labelled models should not be
classifiable, only those with real patterns in data should.

The result is cached to `results/permutation_results.rds`; delete
that file to force recomputation.

```{r permute}
N_PERMS <- 200
PERM_CACHE <- here("results", "permutation_results.rds")

run_permutations <- function(data, features, lrn_id, impute,
                             n_perms, seed) {
  set.seed(seed)
  out <- matrix(NA_real_, nrow = n_perms, ncol = 2,
                dimnames = list(NULL, c("acc", "auc")))
  for (i in seq_len(n_perms)) {
    d_perm <- data
    d_perm$class <- sample(d_perm$class)
    res <- tryCatch(
      run_loocv(d_perm, features, lrn_id, impute),
      error = function(e) list(agg = c(classif.acc = NA_real_,
                                       classif.auc = NA_real_))
    )
    out[i, ] <- c(res$agg["classif.acc"], res$agg["classif.auc"])
  }
  as.data.frame(out)
}

if (file.exists(PERM_CACHE)) {
  perm_results <- readRDS(PERM_CACHE)
  message("Loaded cached permutation results from ", PERM_CACHE)
} else {
  message("Running ", N_PERMS, " permutations per cell (ranger)...")
  t0 <- Sys.time()
  perm_results <- lapply(seq_along(design), function(i) {
    cell <- design[[i]]
    obs <- run_loocv(mat_11, cell$features, "classif.ranger", cell$impute)$agg
    nulls <- run_permutations(mat_11, cell$features, "classif.ranger",
                              cell$impute, N_PERMS, seed = 100 + i)
    list(cell = cell$name, obs = obs, nulls = nulls)
  })
  names(perm_results) <- vapply(design, `[[`, character(1), "name")
  saveRDS(perm_results, PERM_CACHE)
  message("Done in ", round(as.numeric(difftime(Sys.time(), t0, units = "secs"))),
          " s")
}
```

```{r permute-table}
perm_summary <- do.call(rbind, lapply(perm_results, function(r) {
  data.frame(
    cell      = r$cell,
    obs_acc   = unname(r$obs["classif.acc"]),
    obs_auc   = unname(r$obs["classif.auc"]),
    null_acc_mean = mean(r$nulls$acc, na.rm = TRUE),
    null_auc_mean = mean(r$nulls$auc, na.rm = TRUE),
    p_acc     = (1 + sum(r$nulls$acc >= r$obs["classif.acc"], na.rm = TRUE)) /
                (sum(!is.na(r$nulls$acc)) + 1),
    p_auc     = (1 + sum(r$nulls$auc >= r$obs["classif.auc"], na.rm = TRUE)) /
                (sum(!is.na(r$nulls$auc)) + 1)
  )
}))
knitr::kable(perm_summary, digits = 3, row.names = FALSE)
```

```{r permute-plot, fig.width=10, fig.height=6}
nulls_long <- do.call(rbind, lapply(perm_results, function(r) {
  data.frame(cell = r$cell,
             metric = rep(c("acc", "auc"), each = nrow(r$nulls)),
             value  = c(r$nulls$acc, r$nulls$auc))
}))
obs_long <- do.call(rbind, lapply(perm_results, function(r) {
  data.frame(cell = r$cell,
             metric = c("acc", "auc"),
             value  = unname(c(r$obs["classif.acc"], r$obs["classif.auc"])))
}))

ggplot(nulls_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "grey70", colour = "white") +
  geom_vline(data = obs_long, aes(xintercept = value),
             colour = "#ED2939", linewidth = 0.8) +
  facet_grid(cell ~ metric, scales = "free_x") +
  labs(x = "Score under permuted labels (red = observed)", y = "Permutations") +
  theme_minimal(base_size = 10)
```
Indeed thats what we see. Most shuffled-label runs sit around 0.5 on both acc and AUC. The red line is our *observed* score (not a significance cut-off) — the p-value is just the fraction of shuffles that reach or exceed it. For most cells no shuffle matches, so p hits the 1/(200+1) ≈ 0.005 floor; where a handful do, p just inches up. Observed scores this far above the shuffle pile indicate a real pattern.

## 6. Conclusion

The combined data and `ranger` models give a great classification of sensors from helpers using the `SNI` features.

## 7. Session

```{r session}
sessionInfo()
```