---
title: "003 — Data provenance: building a defensible AF3 6mer ground-truth set"
author: "Dan MacLean"
date: "2026-04-24"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  message: false
  warning: false
---

## Goal

This notebook is a **lab-book entry**: it records, in the order it
happened, every step of the data-gathering chain that underpins the
forthcoming QC analysis (notebook 004). Nothing in here computes SNI
or trains a classifier. The aim is that collaborator can read this single file and reproduce, end-to-end:

1. the rubric we pre-registered for grading the oligomeric state of a
   plant NLR;
2. the literature sweep that fed it;
3. the curated table of "known states" we now defend;
4. the PDB IDs we resolved for each entry;
5. the public AF3-prediction archive (Contreras *et al.* 2024) that
   gives us a same-protein-different-stoichiometry test set without
   running AF3 ourselves;
6. the staging conventions that make those archives consumable by the
   `resistosome_pipeline`;
7. the environment and a first sanity-check run of the pipeline on
   Contreras inputs.

Each section names the exact files, scripts, and external IDs
involved. Intermediate decisions (and the judgement calls behind them)
are flagged inline.

## Why we need this

The supervised-learning work in notebooks 001/002 classifies NRC
helpers vs partner sensors on an 18-protein labelled benchmark. That
is **not** the question of scientific interest — it is a within-NRC
proxy for the question of interest, which is:

> When AF3 is given six copies of an NLR, is the resulting hexamer a
> true-positive (matches an experimentally-determined hexamer) or a
> false-positive (the protein's real activated form is not a hexamer
> at all)?

To answer that question we need NLRs whose activated oligomeric state
has been determined experimentally, and AF3 6mer predictions for
those proteins. Sections A–D below build the first; sections G–I
build the second.

---

## A. Pre-registration: the curation rubric

Before pulling any literature we wrote and locked
`data/known_oligomer_states_rubric.md` (v0.1, locked 2026-04-23). The
purpose is to remove cherry-picking discretion from the curator
(future-us, or anyone). Key choices:

- **What we grade**: the *activated / effector-bound* form, because
  that is what AF3 is being asked to predict when given six copies. An
  inactive / autoinhibited structure does **not** set `known_state`.
- **`known_state` vocabulary**: `monomer | dimer | trimer | tetramer
  | pentamer | hexamer | heptamer | octamer_plus | unknown`. There is
  deliberately no `mixed`. If two stoichiometries are reported for one
  protein we call from the higher-resolution / effector-bound /
  more-recent structure; the disagreement is logged in `notes`.
- **State-setting evidence tiers**:
  1. `cryo-EM_highres` — activated complex resolved at ≤4 Å with
     visible protomer count
  2. `cryo-EM_lowres`  — 4–10 Å, protomer count visible
  3. `crystal`         — X-ray of the activated oligomeric complex
- **Non-state-setting evidence tiers** (recorded but never set
  `known_state`): `XL-MS`, `functional_only`, `homology`. Native-MS
  and SEC-MALS are explicitly excluded.
- **Conflicts**: preprints included only if a PDB/EMDB deposit is
  verifiable; otherwise excluded under §G of the rubric.

The same rubric is also applied symmetrically to the existing
18-protein labelled benchmark (§I of the rubric). That regrade
**does not modify** the helper/sensor labels used in notebooks
001/002; it lives alongside them, answering a different question.

A parallel metazoan calibration set is kept under the same rubric in
`data/known_oligomer_states_metazoan.csv` (planned; metazoan triage is
still pending — see §M of this notebook).

## B. PubMed sweep — five pre-registered queries

Five queries were committed under §K of the rubric on 2026-04-23. Any
query added after that date is tagged `post_hoc` in the audit log.

| Tag | Scope | Hits |
|----:|:------|----:|
| **P1** | plant NLR × structural method (`cryo-EM`/`crystal`/`X-ray structure`) | 37 |
| **P2** | resistosome-centric (`"resistosome"` and variants) | 102 |
| **P3** | named-gene sweep (curated list of 13 plant NLRs we already expect structures for; ZAR1, Sr35, RPP1, etc.) | 101 |
| **M1** | metazoan calibration — NAIP/NLRC4 only (NLRP3 deliberately excluded) | 205 |
| **M2** | metazoan calibration — apoptosome family | 153 |

Total 598 raw, **536 unique** PMIDs after dedup. Honesty notes: P3 is
a curation decision, not a derivation; that bias is acknowledged in
the rubric so it can be challenged.

The mechanical pipeline (re-runnable from `data/pubmed_raw/triage.py`)
is roughly:

```python
# Pseudocode — see data/pubmed_raw/triage.py for the real thing.
for query in [P1, P2, P3, M1, M2]:
    pmids   = curl(esearch, query)              # → P{1..5}_esearch.json
    summary = curl(esummary, pmids, batches=5)  # → esummary_*.json
    abstract = curl(efetch,  pmids, batches=5)  # → efetch_*.xml
unique = sorted(set.union(*query_pmids))         # → unique_pmids.json (536)
abstracts = parse_titles_abstracts_pubtypes_etc()# → abstracts.json
```

## C. Triage — mechanical first pass + human survivor pass

Each unique PMID is dispositioned by `triage.py` using a closed set of
9 categories:

`included | not_plant_NLR | homology_only | no_structure | inactive_only | review | duplicate | ambiguous | needs_human_review`

Mechanical rules read titles + abstracts + publication types; rows
that fail every rule are tagged `needs_human_review`. The 20 strongest
plant hits (high P1+P2+P3 overlap) were then reviewed by hand and
written into the audit log columns `final_disposition` and
`manual_notes`. Outputs:

- `data/known_oligomer_states_audit.csv` — one row per unique PMID,
  full provenance.
- `data/pubmed_raw/abstracts.json` — full title/abstract/pubtypes
  cache for re-grading.

After the manual pass the 8 plant hits that survived as
state-setting structural papers were promoted into the curated table
in §D. **Two sets of work are deferred**: 97 weak-plant survivors and
287 metazoan PMIDs are still in `ambiguous` and have not yet been
human-reviewed.

## D. The curated literature table — 8 entries

```{r setup, include=FALSE}
library(here)
library(dplyr)
library(knitr)
```

```{r curated-table}
known <- read.csv(here("data", "known_oligomer_states.csv"),
                  stringsAsFactors = FALSE)

literature <- known |>
  filter(in_current_18 == "FALSE") |>
  select(ID, protein_family, organism, known_state, evidence_tier,
         source_pmid, source_pdb, usable_for_analysis)

kable(literature, caption = "Literature plant NLR entries (8). All graded under v0.1 rubric.")
```

Three of these required explicit judgement calls; they are flagged
inline in the `notes` column of the full CSV:

- **MLA13** (PMID 39948409, EMBO J 2025). The published cryo-EM is a
  1:1 effector heterodimer. We did **not** call this `monomer`. The
  rubric says: absence of a resistosome in this purification is not
  evidence that the protein cannot form one in another condition.
  Hence `known_state = unknown`, `evidence_tier = cryo-EM_highres`,
  with the heterodimer logged in `notes`.
- **NRC4 preprint** (PMID 38187616, bioRxiv) was **excluded** under
  rubric §G — preprints are only included if a PDB/EMDB deposit is
  verifiable, and this one's deposit was not at the time of curation.
- **RPS2** (PMID 41864205, *Cell* 2026, alongside WAI3): the abstract
  reports an octameric RPS2 complex but no separate RPS2-only PDB
  deposit surfaced via NCBI elink at curation time. Recorded with
  `source_pdb` blank and a flagged `notes` entry.

## E. Resolving PDB IDs

Initial attempt: query the RCSB search API directly for entries
citing a given PMID via
`rcsb_primary_citation.pdbx_database_id_PubMed`. This returned
`HTTP 400 — search is not enabled on attribute`. The same was true of
`citation.pdbx_database_id_PubMed`. The RCSB JSON-search interface
does not expose PMID as a searchable citation field.

The working pivot is:

```bash
# Pseudocode — for each curated PMID:
elink   db=pubmed dbfrom=pubmed db=structure id=$PMID  # → MMDB UIDs
esummary db=structure id=$mmdb_uids                    # → PDB codes (PdbAcc)
```

Resolved (full IDs in `data/pdb_lookup/mmdb_to_pdb.json`):

| Protein | PDB IDs |
|:--------|:--------|
| AtZAR1  | 6J5T, 6J6I |
| Sr35    | 7XC2, 7XX2, 7XVG, 7XE0, 7XDS |
| RPP1    | 7CRC, 7CRB, 7DFV |
| ROQ1    | 7JLX, 7JLV, 7JLU |
| NbNRC2  | 9FP6 (active), 8RFH (resting) |
| WAI3    | 9H73, 9H4I, 9H2L (RPS2 likely grouped here) |
| MLA13   | 9FYC |

## F. Symmetric regrade of the 18-protein benchmark

The same rubric was applied to the 18 proteins used in
notebooks 001/002. Most stay `known_state = unknown` (functional
helper/sensor classification ≠ structural evidence). Two NRC entries
inherit the NbNRC2 hexamer call:

```{r in18}
in18 <- known |>
  filter(in_current_18 == "TRUE") |>
  count(class = known_state, name = "n")
kable(in18, caption = "Regraded 18-protein benchmark by known_state.")
```

Two entries (`cnl_11990`, `rpa1`) were flagged
`usable_for_analysis = FALSE` per Dan's call: identity unverified
during curation, so they are recorded but excluded from any downstream
analysis. The notebook-002 helper/sensor labels are **not** modified.

## G. Looking for AF3 predictions in public databases

We scoped before generating: is there a public archive of AF3
multimer predictions we could use without running AF3 ourselves?

- **AlphaFold DB** (alphafold.ebi.ac.uk): AF2 monomers only. Not
  useful for the 6mer question.
- **AF3 Server outputs**: not centrally archived; per-paper.
- **Per-paper supplementary on Zenodo**: yes, in some cases.

The hit was Contreras *et al.* 2024 (Sci Adv, PMID 39504373, DOI
`10.1126/sciadv.adr2594`) — the same paper that defines our NbNRC2
hexamer ground truth. Their GitHub
(`github.com/amiralito/NRC2Hexamer`) points at two Zenodo deposits;
the second one is the relevant archive of AF3 outputs.

## H. The Contreras Zenodo (10.5281/zenodo.11546022)

17.1 GB across 11 files. The metadata workbook `Data S1.xlsx` (also
copied to `data/contreras_af3/Data_S1.xlsx`) lists 29 NLRs across 5
sheets. Pulled via `data/contreras_af3/fetch.sh`:

```bash
# Each file of the deposit, resumable, into data/contreras_af3/zips/
for f in NRC2_{tetra,penta,hexa,hepta,octa}mers.zip \
         Benchmarks.zip AF_benchmark.zip CC.zip CCG10.zip CCR.zip; do
    curl -L --fail --retry 3 -C - \
         -o zips/$f \
         "https://zenodo.org/records/11546022/files/$f"
done
```

The five sheets of `Data_S1.xlsx` map onto the archives as follows:

| Sheet | Content | In Zenodo? |
|---:|:----|:----|
| 1. AlphaFold benchmark | NbNRC2a × {AF3+lipids, AF3, AF2} | partial — AF3 only |
| 2. NRC2 benchmark | NbNRC2a × **{4,5,6,7,8}-mer × 10 seeds** = 50 models | yes — `NRC2_*mers.zip` |
| 3. NRCs | 11 NRC helpers × {5,6}-mer | **NO** — metrics reported, structures not deposited |
| 4. cryo-EM benchmark | NbNRC2a / AtZAR1 / Sr35 × {5,6}-mer | yes — `Benchmarks.zip` |
| 5. CC-types NLRs | 8 CC-NLR + 6 CCG10-NLR + 2 CCR-NLR × {5,6}-mer | yes — `CC.zip`, `CCG10.zip`, `CCR.zip` |

So the actually-usable inventory is **89 prediction entries**: 50 of
the NbNRC2a stoichiometry sweep, 6 cryo-EM benchmark, 32 CC-types,
plus one duplicate hexamer in `AF_benchmark.zip`.

### Coverage of our literature ground truth

Cross-referencing §D against Contreras' panel:

| GT protein | GT state | In Contreras? | Stoichs predicted | Use |
|:--|:--|:-:|:-:|:--|
| AtZAR1 | pentamer | ✅ | {5, 6} | TP @ 5, FP @ 6 |
| Sr35 | pentamer | ✅ | {5, 6} | TP @ 5, FP @ 6 |
| **NbNRC2** | **hexamer** | ✅ | **{4, 5, 6, 7, 8}** × 10 seeds | TP @ 6 plus FP at every other stoich; the gold dataset |
| RPS2 | octamer⁺ | ✅ | {5, 6} | both wrong-stoich; "AF3 underestimates" only |
| RPP1 | tetramer | ❌ | — | gap (TNL, out of Contreras scope) |
| ROQ1 | tetramer | ❌ | — | gap (TNL) |
| MLA13 | unknown | ❌ | — | gap (barley CNL) |
| WAI3 | octamer⁺ | ❌ | — | gap (wheat CCG10) |

**Decision**: we are **not** running AF3 ourselves on the four gap
proteins. AF3 Server jobs at this stoichiometry size are long, prone
to queue failures, and would be an open-ended commitment. The QC
notebook will be scoped to what Contreras gives us; the gaps are
documented here as known limitations.

## I. Staging Contreras into the resistosome pipeline

Each Contreras AF3 prediction is a directory of ~5–170 MB containing:

```
fold_<protein>_..._<seed-token>/
    <stem>_model_0.cif                      # the top-ranked model
    <stem>_summary_confidences_0.json       # ipTM, pTM, chain pTM
    <stem>_full_data_0.json                 # ~170 MB, atomic confidences
    <stem>_model_{1,2,3,4}.cif              # alternate samples
    <stem>_summary_confidences_{1,2,3,4}.json
    <stem>_job_request.json
```

Three things had to be normalised before the
`resistosome_pipeline` could ingest them:

1. **Three different seed-token conventions** are used across the
   archives:
   - `_seed_random_<N>` (most files; seeds 2..10 of the NRC2 sweep)
   - `_seed_<N>` (NRC2 tetramers, seed 1)
   - `_<YYYY>_<MM>_<DD>_<HH>_<MM>` timestamp (the "seed 1" run in most
     archives; recorded as Seed=1 in the manifest)
2. **Each seed directory contains 5 candidate AF3 models**
   (`model_0..model_4`). Contreras' reported stats are on `model_0`
   only, so we keep `model_0` and discard the rest.
3. The pipeline's seed-grouping regex is `_seed(\d+)`, which matches
   `_seed_<N>` but not `_seed_random_<N>` or the timestamp form.

The staging script `data/contreras_af3/stage.py` handles all three:

```python
# Pseudocode — see stage.py for the real regexes.
for zip in zips/*.zip:
    for member in zip:
        skip __MACOSX/, ._*, full_data_*, model_{1..4}, *_seed/    # cruft
        match member against the 3 seed-token patterns above
        emit:  staging/<archive>/<base>_seed<N>/
                  <base>_seed<N>_model.cif
                  <base>_seed<N>_summary_confidences.json
```

This produced **89 (model.cif, summary_confidences.json) pairs** —
matches the manifest exactly:

```
staging/AF_benchmark/         1 entry  (NbNRC2a hexamer, AF3)
staging/Benchmarks/           6 entries (3 proteins × 2 stoichs)
staging/CC/                  16 entries (8 proteins × 2 stoichs)
staging/CCG10/               12 entries (6 proteins × 2 stoichs)
staging/CCR/                  4 entries (2 proteins × 2 stoichs)
staging/NRC2_{tetra..octa}mers/ 10 each = 50 entries (NbNRC2a × 5 stoichs × 10 seeds)
```

**Triple-deposited NbNRC2a hexamer.** The same `seed_random_1` /
timestamp run for NbNRC2a hexamer appears in `AF_benchmark.zip`,
`Benchmarks.zip` (as `nbnrc2a_ccnbarc_oleic_seed1`), and as `seed1` of
`NRC2_hexamers.zip`. They are the same model. Notebook 004 will pick
one canonical copy and dedupe before computing the feature matrix.

## J. SNI pipeline environment and a first run

**Environment** (`mamba env list` → `sni-pipeline`):

```bash
mamba create -n sni-pipeline \
    -c https://repo.prefix.dev/conda-forge \
    -c https://repo.prefix.dev/bioconda \
    python=3.11 biopython=1.84 freesasa=2.2.1 numpy=1.26.3 \
    scipy=1.15.2 pandas=2.3.3 openpyxl=3.1.5 dssp hmmer
```

(Channel URLs go through `repo.prefix.dev` for licence-compliance
reasons; do not use the bare `-c conda-forge` form even though the
local `~/.condarc` happens to alias it through.)

**Trial run** on the 6 cryo-EM benchmark predictions (NbNRC2a /
ZAR1 / Sr35 × {penta, hexa}):

```bash
cd SolNRCH_foldome/resistosome_pipeline
mamba run -n sni-pipeline python run_pipeline.py \
    --input_dir  sni_ml/data/contreras_af3/staging/Benchmarks \
    --output_dir /tmp/sni_test_benchmarks \
    --hmm        SolNRCH_foldome/structural_analysis/main/MADA.hmm \
    --hmm_fallback_dssp \
    --workers 4
```

Six structures → 16 s wallclock. All 11 canonical SNI features
(`ipTM_LCB`, `Sum_CONTACTS`, `BSA_INT_PROTO`, `SD_THETA_ROT`,
`S_PROTO`, `D_APEX`, `THETA_APEX_mean`, `L_APEX_mean`, `H_ABS_mean`,
`MU_H_mean`, `D_MHD_P_mean`) computed, save for `D_MHD_P_mean`, which
needs MHD/P-loop annotation CSVs we have only for the NRC network.

First numbers (full table will be re-derived in 004; this is
qualitative reassurance that the wiring is right):

| ID | ipTM_LCB | Sum_CONTACTS | SD_θ_rot | S_proto | D_apex |
|:---|---:|---:|---:|---:|---:|
| **zar1 penta** ✅ | 0.63 | 299 | **0.07** | **0.03** | 12.6 |
| zar1 hexa ❌ | 0.61 | 152 | **14.1** | **1.45** | 14.9 |
| **nbnrc2a hexa** ✅ | 0.71 | 187 | 0.13 | 0.05 | 21.0 |
| nbnrc2a penta ❌ | 0.66 | 186 | 0.17 | 0.07 | 18.7 |
| **sr35 penta** ✅ | 0.64 | 208 | 0.10 | 0.07 | 20.1 |
| sr35 hexa ❌ | 0.64 | 209 | 0.17 | 0.07 | 21.4 |

Read: ZAR1 is exactly the SNI behaviour you'd hope for — a
wrong-stoich AF3 hexamer has 200× the rotational asymmetry of the
correct pentamer. NbNRC2a is mid-strength. **Sr35 is a clean
non-discrimination** — AF3 builds geometrically clean pentamers *and*
hexamers, and SNI cannot tell them apart. That single-protein
counter-example is the kind of signal the QC analysis in 004 needs
to characterise across the full 89-entry panel.

## K. What this notebook does not do

- It does not compute the full 89-entry feature matrix.
- It does not train, evaluate, or compare any classifier.
- It does not regrade the 002 helper/sensor labels.
- It does not propose AF3 server runs for the gap proteins.

All of those belong in notebook 004 (planned).

## L. Open follow-ups

- **`D_MHD_P_mean` annotation gap** for non-NRC Contreras proteins
  (8 CC-NLRs, 6 CCG10-NLRs, 2 CCR-NLRs, ZAR1, Sr35). The NRC-network
  CSVs at `SolNRCH_foldome/structural_analysis/main/*_coords.csv` do
  not cover them. Either an HMM-driven motif-coord scan, or accept
  10/11 features for those rows, or drop the metric from the QC
  feature set.
- **Sr35 missed the MADA HMM** in the trial run and fell back to
  DSSP for α1, which downstreams to `THETA_APEX_mean`, `L_APEX_mean`,
  `MU_H_mean`. Worth checking whether the existing 18-protein
  benchmark hit HMM for Sr35 or also fell back, and what the
  consequences are.
- **Triple-deposited NbNRC2a hexamer**: pick a canonical copy in 004.
- **Weak-plant scan**: 97 PMIDs still tagged `ambiguous` in the audit
  log have not had a human pass.
- **Metazoan calibration**: 287 PMIDs across M1+M2 still `ambiguous`.
  None of the M-set has been promoted into a curated table yet.
- **RPS2 PDB deposit**: re-check NCBI elink in a few weeks; the *Cell*
  2026 deposits may not yet be fully indexed.

## M. File index

Everything created in this lineage, with one-line descriptions:

| Path (relative to `sni_ml/`) | What |
|:--|:--|
| `data/known_oligomer_states_rubric.md` | v0.1 pre-registered rubric (locked 2026-04-23) |
| `data/known_oligomer_states.csv` | curated table — 26 rows (18 benchmark + 8 literature) |
| `data/known_oligomer_states_audit.csv` | per-PMID disposition log, 536 rows |
| `data/pubmed_raw/` | raw esearch / esummary / efetch outputs and `triage.py` |
| `data/pdb_lookup/` | NCBI elink → MMDB → PDB resolution outputs |
| `data/contreras_af3/Data_S1.xlsx` | Contreras *et al.* 2024 supplementary metadata |
| `data/contreras_af3/manifest/*.csv` | each sheet of `Data_S1.xlsx` as a CSV |
| `data/contreras_af3/fetch.sh` | resumable Zenodo downloader for the 10 archives |
| `data/contreras_af3/zips/` | original 14.4 GB of Zenodo archives (kept for re-staging) |
| `data/contreras_af3/stage.py` | seed-convention normaliser; produces `staging/` |
| `data/contreras_af3/staging/` | 89 (model.cif, summary_confidences.json) pairs |

External resources cited:

- Contreras *et al.* 2024, *Sci Adv* 10, eadr2594. PMID 39504373, DOI
  `10.1126/sciadv.adr2594`.
- Contreras *et al.* 2024, AF3-predicted NLR resistosome structures.
  Zenodo: `10.5281/zenodo.11546022`.
- `amiralito/NRC2Hexamer` (GitHub).
- NCBI E-utilities (esearch, esummary, efetch, elink) and MMDB/Structure.
- RCSB PDB (REST/JSON search API) — failed for the citation→entry
  lookup; recorded for the next person who tries it.
- The `resistosome_pipeline` lives at
  `SolNRCH_foldome/resistosome_pipeline/`.
