---
title: "004 — QC of the AF3 hexamer call: Contreras feature matrix vs ground truth"
author: "Dan MacLean"
date: "2026-04-24"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  message: false
  warning: false
---

```{r setup, include=FALSE}
library(here)
library(dplyr)
library(tidyr)
library(knitr)
fm_path <- here("results", "004_full_run", "feature_matrix.csv")
fm <- read.csv(fm_path, stringsAsFactors = FALSE, na.strings = c("", "NA"))
# pandas writes booleans as "True"/"False" strings — coerce back
fm$is_triple_deposit <- fm$is_triple_deposit == "True"
fm$call_correct      <- ifelse(fm$call_correct == "True", TRUE,
                        ifelse(fm$call_correct == "False", FALSE, NA))
```

## Aims

1. **Compute** the 11 canonical SNI (Structural Novelty Index)
   features for every plant-NLR AF3 model deposited by Contreras *et
   al.* 2024 on Zenodo (89 model files in total).
2. **Join** those features to a small set of plant NLRs whose
   activated oligomeric state has been determined experimentally
   (cryo-EM, mostly).
3. **Show** how SNI separates a correct AF3 oligomer call from an
   incorrect one in the cases where we can check it directly.
4. **Surface** any feature-coverage or naming gaps that would bias a
   downstream classifier (notebook 005).

## Rationale

The medium-term goal is a classifier that, given an AF3 prediction of
a plant CC-NLR pulled at 6 copies, says *“this is a real hexameric
resistosome”* or *“this is a forced hexamer that AF3 has been talked
into modelling”* (a 6mer **false positive**, 6mer-FP). The
classifier features are the 11 SNI metrics — symmetry RMSDs,
contact counts, interface area, α1-helix geometry, MHD–P-loop
distance, etc. Before training anything, we need to verify that:

- The pipeline runs on every Contreras model without falling over.
- For the (small) subset of proteins where we *know* the right answer
  from cryo-EM, the canonical SNI features point in the expected
  direction when AF3 is forced to model the wrong stoichiometry.
- We know which features are missing on which proteins, and why,
  before writing classifier code that will silently drop rows.

This notebook is the experimental record of that verification.

## Materials

### What Contreras *et al.* deposited

Contreras *et al.* 2024 (*Sci Adv* 10, eadr2594; PMID 39504373; data
on Zenodo `10.5281/zenodo.11546022`) ran AlphaFold-Multimer 3 on a
panel of plant NLRs, forcing the model to assemble homo-oligomers at
fixed stoichiometries. Their deposit is a set of 10 zip files we
will refer to as **archives** throughout this notebook. Each archive
groups predictions by purpose:

| Archive | What's in it | # protein × stoichiometry rows |
|:--|:--|:--|
| `AF_benchmark/` | one AF3 prediction of NbNRC2a hexamer (deposit benchmark) | 1 |
| `Benchmarks/` | 3 reference proteins (ZAR1, Sr35, NbNRC2a) at hexa **and** penta — the small "does the pipeline behave?" set | 6 |
| `CC/`, `CCG10/`, `CCR/` | broader CC-NLR panel — each protein modelled at hexa **and** penta | 16 / 12 / 4 |
| `NRC2_tetramers/` … `NRC2_octamers/` | NbNRC2a forced to assemble at 4, 5, 6, 7 and 8 copies; **10 random AF3 seeds per stoichiometry** — the "gold sweep" | 5 × 1 |

Each protein × stoichiometry × seed prediction lives as a directory
of AF3 outputs (5 candidate models, summary JSON, full PAE/PDE
arrays). In notebook 003 we extracted only the top-ranked model per
seed (`*_model_0.cif` plus `*_summary_confidences_0.json`),
dropped the 4 alternative models and the bulky raw arrays, normalised
the seed-naming, and laid the result out at
`sni_ml/data/contreras_af3/staging/<archive>/<protein>_seed<N>/`. The
full file count is **89 (model.cif, summary_confidences.json) pairs**
across the 10 archives.

A **stoichiometry** here means the number of identical NLR copies
AlphaFold-Multimer was asked to model into the same complex. We refer
to them as 4-mer, 5-mer (= pentamer), 6-mer (= hexamer), 7-mer,
8-mer.

### The ground-truth panel

Notebook 003 produced a curated literature table at
`sni_ml/data/known_oligomer_states.csv` — 26 rows, of which 8 are
plant NLRs with a published activated-form structure resolved at
cryo-EM tier-1–3 evidence. Of those 8, **four** also appear in the
Contreras Zenodo deposit, and so can be cross-checked here:

| Protein | Activated state | Source | Contreras coverage |
|:--|:--|:--|:--|
| ZAR1 (Arabidopsis) | pentamer | Wang *et al.* 2019, PDB 6J5T | hexa + penta in `Benchmarks/` |
| Sr35 (wheat) | pentamer | Zhao *et al.* 2022, PDB 7XC2 | hexa + penta in `Benchmarks/` |
| NbNRC2 / nbnrc2a (*N. benthamiana*) | hexamer | Madhuprakash *et al.* 2024, PDB 9FP6 | full 4–8 sweep + Benchmarks + AF_benchmark |
| RPS2 (Arabidopsis) | octamer (≥8) | *Cell* 2026 | hexa + penta in `CCG10/` (no octa modelled) |

The remaining 4 ground-truth proteins (RPP1, ROQ1, MLA13, WAI3) are
not in the Contreras deposit and are out of scope here.

### The 11 SNI features

The pipeline's canonical feature set (the 11 we will be looking at
throughout):

- `ipTM_LCB` — AF3 interface-pTM, lower 95 % confidence bound across
  seeds. *Confidence in the interface AF3 modelled.*
- `Sum_CONTACTS` — total inter-chain heavy-atom contacts at 3.5 Å.
  *How densely chains touch.*
- `BSA_INT_PROTO` — buried surface area of the protomer interface.
- `SD_THETA_ROT` — standard deviation of inter-chain rotation angle.
  *0 for a perfect ring; ≫0 when chains splay.*
- `S_PROTO` — protomer-symmetry RMSD across the ring.
- `D_APEX` — radial distance from ring axis to the α1 apex.
- `THETA_APEX_mean` — mean apex tilt (α1 angle).
- `L_APEX_mean` — mean α1 length.
- `H_ABS_mean` — mean α1 hydrophobicity.
- `MU_H_mean` — mean α1 amphipathic moment.
- `D_MHD_P_mean` — mean intra-protomer distance from the MHD motif
  to the P-loop. *Requires per-protein motif-coordinate annotations.*

## Methods

### Toolchain audit

Before running anything, we verified the external tools the pipeline
shells out to. The pipeline source (`resistosome/hmm.py`,
`resistosome/helix.py`) calls `subprocess.run(['hmmsearch', …])`
and `subprocess.run(['mkdssp', …])` — i.e. it relies on PATH lookup
for both binaries. The matrix below is what was confirmed before
running:

| Tool | Purpose | Resolved binary | Version |
|:--|:--|:--|:--|
| `hmmsearch` | MADA HMM search → α1 boundaries | `…/envs/sni-pipeline/bin/hmmsearch` | HMMER 3.4 |
| `mkdssp` | secondary-structure for α1 fallback | `…/envs/sni-pipeline/bin/mkdssp` | 4.6.1 |
| `python` | pipeline interpreter | `…/envs/sni-pipeline/bin/python` | 3.11 |

Python deps (`biopython`, `freesasa`, `numpy`, `scipy`, `pandas`,
`openpyxl`) all install-verified in the same env.

The driver script `sni_ml/data/contreras_af3/run_pipeline_all.sh`
prepends `…/envs/sni-pipeline/bin` to PATH, then runs `command -v
hmmsearch mkdssp` as a hard precondition (`exit 2` on miss) before
launching any pipeline call. This guards against the failure mode
recorded in the incident below.

#### Incident: the first run silently fell back to DSSP everywhere

The first pipeline invocation in this notebook was launched as
`/Users/macleand/mamba/envs/sni-pipeline/bin/python run_pipeline.py …`.
That picks up the env's Python and its installed packages, but does
**not** add the env's `bin/` to PATH — so `subprocess.run(['hmmsearch',
…])` from inside the pipeline got *command not found* on every call.
The pipeline silently fell back to DSSP for α1 boundaries on every
protein in every archive (the fallback is `--hmm_fallback_dssp`,
which is on). The α1-derived features
(`THETA_APEX_mean`, `L_APEX_mean`, `MU_H_mean`) were therefore
DSSP-derived in every row of the first matrix, not MADA-derived.

The discrepancy between HMM and DSSP α1 boundaries is non-trivial
for proteins whose α1 includes a defined MADA motif at the very
N-terminus: for NbNRC2a hexamer, `THETA_APEX_mean` changed from
≈18° (DSSP) to ≈47° (HMM), and `L_APEX_mean` changed from ≈51
residues (DSSP, picking up the entire helical run) to 21 residues
(HMM, restricted to the MADA motif). A side-by-side diff is at
`results/004_full_run/feature_matrix_dsspfallback.csv` (frozen
copy of the first run) and `results/004_full_run/feature_matrix.csv`
(corrected, current). Critically, the **non-α1 separators**
(`ipTM_LCB`, `Sum_CONTACTS`, `BSA_INT_PROTO`, `SD_THETA_ROT`,
`S_PROTO`, `D_APEX`) are bit-identical between the two runs — they
do not depend on α1 boundaries — so the headline TP/FP findings in
R3–R4 below were unaffected. Lesson recorded in the driver as the
mandatory tool-presence check; the bad matrix is preserved as a
diff target rather than deleted.

### Pipeline run

The driver script
`sni_ml/data/contreras_af3/run_pipeline_all.sh` calls
`SolNRCH_foldome/resistosome_pipeline/run_pipeline.py` once per
archive. Running per-archive (rather than over a single flat input
directory) avoids filename collisions on
`nbnrc2a_..._seed1`, which appears in three archives
(`AF_benchmark`, `Benchmarks`, `NRC2_hexamers`).

```
export PATH=/Users/macleand/mamba/envs/sni-pipeline/bin:$PATH
command -v hmmsearch mkdssp || exit 2

for arc in AF_benchmark Benchmarks CC CCG10 CCR \
           NRC2_tetramers NRC2_pentamers NRC2_hexamers \
           NRC2_heptamers NRC2_octamers:
    python run_pipeline.py \
        --input_dir  staging/<arc>/ \
        --output_dir results/004_full_run/<arc>/ \
        --mhd_csv    .../NRCH_MHD_nlrexp_coords.csv \
        --ploop_csv  .../NRCH_ploop_nlrexp_coords.csv \
        --nbarc_csv  .../NRCH_clade_filtered_..._coords.csv \
        --hmm        .../MADA.hmm \
        --hmm_fallback_dssp \
        --workers 8 \
        --log_level WARNING
```

The pipeline groups input directories by stripping the `_seed<N>`
suffix from the directory name. The result is one row per
`(protein, predicted_oligomer)` per archive, with `N_replicates`
equal to the number of seeds in that group. The NRC2_*mers archives
each carry `N_replicates = 10`; everything else `N_replicates = 1`.

Per-archive runtime was a few seconds; full sweep ≈ 2 min wall-clock
on this machine. Logs are at
`results/004_full_run/master.log` (driver) and
`results/004_full_run/<archive>/run.log` (pipeline stderr).

### Building the feature matrix

The merge step
`sni_ml/data/contreras_af3/build_feature_matrix.py`:

1. Concatenates the 10 per-archive `resistosome_analysis_summary.csv`
   files (58 columns each), prepending an `archive` column.
2. Parses the pipeline's `ID` field to recover `protein` (lowercased,
   stripped of the `_ccnbarc / _<oligo-token> / _oleic` decoration
   Contreras attached) and `predicted_oligomer` (`tetra→4 …
   octa→8`; no token = Contreras default = 6).
3. Joins to `data/known_oligomer_states.csv` on lowercased `protein`
   to attach `gt_oligomer`, evidence tier, and other metadata.
4. Adds `call_correct = (predicted_oligomer == gt_oligomer)` for
   rows where ground truth is known; `NA` otherwise.
5. Adds `is_triple_deposit = TRUE` for the three rows that hold the
   same NbNRC2a hexamer model under different archive provenance
   (`AF_benchmark`, `Benchmarks`, `NRC2_hexamers`).

The output is `results/004_full_run/feature_matrix.csv` (also
`.xlsx`). Every result and table below is computed from that one
file.

## Results

### R1. Matrix shape and ground-truth coverage

```{r matrix-shape}
cat("rows:", nrow(fm), "  cols:", ncol(fm), "\n")
cat("rows with GT match:", sum(!is.na(fm$gt_oligomer)), "\n")
cat("rows w/o  GT match:", sum( is.na(fm$gt_oligomer)), "\n")
cat("triple-deposit rows:", sum(fm$is_triple_deposit), "\n")
```

44 rows from 89 staged models — the difference is the seed
collapsing the pipeline does inside the NRC2 sweep. 14 rows have a
joined ground-truth state; the remaining 30 are CC/CCG10/CCR
proteins and a handful of others where we have no published
activated structure.

### R2. What is in the matrix, by archive

```{r matrix-layout}
fm |>
  mutate(stoich = recode(predicted_oligomer,
                         `4`="tetra", `5`="penta", `6`="hexa",
                         `7`="hepta", `8`="octa")) |>
  group_by(archive, protein) |>
  summarise(stoichiometries = paste(sort(unique(stoich)), collapse=", "),
            n_seeds_total   = sum(N_replicates),
            .groups = "drop") |>
  arrange(archive, protein) |>
  kable()
```

The CC, CCG10 and CCR archives each cover their proteins at hexa
**and** penta, which means we can ask the
"forced-wrong-stoichiometry" question of every one of them — even
the ones with no ground truth — because at most one of the two
calls can be right and the SNI features should reflect that.

### R3. Hexamer call against ground truth

For the 14 GT-bearing rows, compare the predicted stoichiometry to
the cryo-EM-determined oligomer and look at the four most
discriminating SNI features. **TP** = AF3 was forced at the right
stoichiometry; **FP** = AF3 was forced at a wrong one.

```{r call-table}
fm |>
  filter(!is.na(gt_oligomer)) |>
  mutate(call = ifelse(call_correct, "TP", "FP")) |>
  select(archive, protein, predicted_oligomer, gt_oligomer,
         N_replicates, ipTM_LCB, Sum_CONTACTS, SD_THETA_ROT, S_PROTO,
         is_triple_deposit, call) |>
  arrange(protein, archive, predicted_oligomer) |>
  kable(digits = 3)
```

Reading the four candidate-separator columns by eye, with no
statistical test applied:

- **ZAR1**: penta (TP) vs hexa (FP) — `SD_THETA_ROT` 0.07 vs 14.1,
  `S_PROTO` 0.03 vs 1.45, `Sum_CONTACTS` 299 vs 152. Numerically
  large gap on each.
- **NbNRC2a**: hexa (TP, n=10 seeds) vs penta (FP, n=1 in
  Benchmarks) — `SD_THETA_ROT` 0.13 vs 0.17, `S_PROTO` 0.05 vs
  0.07. Numerically small gap, in the same direction as ZAR1.
- **Sr35**: penta (TP) vs hexa (FP) — `SD_THETA_ROT` 0.10 vs 0.17,
  `S_PROTO` 0.07 vs 0.07. Small gap on `SD_THETA_ROT`, no gap on
  `S_PROTO`.
- **RPS2**: both rows are forced FPs (the panel only modelled hexa
  and penta, neither matches the GT octamer); we record the
  feature values for future comparison.

These are observations from the table only. Whether the gaps are
robust enough to discriminate TP from FP across proteins is a
question for 005's EDA, not for this notebook.

### R4. The NbNRC2a stoichiometry sweep — the gold panel

The `NRC2_tetramers/` … `NRC2_octamers/` archives are the single
most informative dataset in this project: the same protein, ground
truth = 6, AF3 forced at 4, 5, 6, 7 and 8, each at 10 random seeds.
Anything that separates 6 from {4, 5, 7, 8} *within* this protein is
a candidate 6mer-FP signal.

```{r nrc-sweep}
fm |>
  filter(grepl("^NRC2_", archive)) |>
  arrange(predicted_oligomer) |>
  select(archive, predicted_oligomer, N_replicates,
         ipTM_LCB, Sum_CONTACTS, BSA_INT_PROTO,
         SD_THETA_ROT, S_PROTO, D_APEX,
         THETA_APEX_mean, L_APEX_mean,
         H_ABS_mean, MU_H_mean, D_MHD_P_mean) |>
  kable(digits = 3,
        caption = "NbNRC2a 4/5/6/7/8-mer × 10 seeds, all 11 SNI features.")
```

### R5. The triple-deposited NbNRC2a hexamer

The same NbNRC2a hexamer model_0 (seed 1) appears in three places:
the `AF_benchmark/` archive (deposit-level benchmark), the
`Benchmarks/` archive (paper-level benchmark), and as seed 1 of the
10-seed `NRC2_hexamers/` sweep. The matrix carries all three rows.

```{r triple}
fm |>
  filter(is_triple_deposit) |>
  select(archive, protein, predicted_oligomer, N_replicates,
         ipTM_LCB, Sum_CONTACTS, BSA_INT_PROTO,
         SD_THETA_ROT, S_PROTO, D_APEX) |>
  kable(digits = 3)
```

The `NRC2_hexamers` row is the canonical aggregate (10 seeds, with
`_sd` columns reflecting between-seed variance). The other two are
single-seed re-runs deposited for benchmarking.

**Decision applied in 005:** keep the `NRC2_hexamers` row for any
across-protein analysis (it is the only one with replicate variance
information); drop the `AF_benchmark` and `Benchmarks` hexamer rows
from classifier training so NbNRC2a is not triple-counted. The two
duplicates remain in the matrix with `is_triple_deposit = TRUE` so
they can be inspected for deposit-internal consistency.

### R6. Feature coverage and which proteins fell back to DSSP

```{r coverage}
SNI <- c("ipTM_LCB","Sum_CONTACTS","BSA_INT_PROTO","SD_THETA_ROT",
         "S_PROTO","D_APEX","THETA_APEX_mean","L_APEX_mean",
         "H_ABS_mean","MU_H_mean","D_MHD_P_mean")
fm |>
  select(archive, protein, predicted_oligomer, all_of(SNI)) |>
  pivot_longer(all_of(SNI), names_to = "feature", values_to = "v") |>
  group_by(feature) |>
  summarise(n_present = sum(!is.na(v)),
            n_missing = sum(is.na(v)),
            .groups   = "drop") |>
  arrange(desc(n_missing)) |>
  kable(caption = "Per-feature presence across the 44 rows.")
```

Two coverage issues to flag because they will bias any downstream
model if not addressed:

- **`D_MHD_P_mean`** is `NA` on every row that is not in the
  NRC-network panel. The pipeline computes this distance from
  per-protein motif-coordinate CSVs (`--mhd_csv`, `--ploop_csv`),
  which currently cover only NRC-family proteins. The CC, CCG10,
  CCR, ZAR1 and Sr35 rows therefore have a hole in feature 11.
- **MADA HMM hit/miss is per-protein.** With `hmmsearch` correctly
  on PATH (see *Methods → Incident*), the MADA HMM still misses some
  proteins on biological grounds — those proteins' α1 is then defined
  by the DSSP fallback, which is less precise. The `hmm/hmmsearch_domtbl.out`
  artefact in each archive directory records which proteins hit:

```{r hmm-per-protein}
arc_dirs <- list.files(here("results", "004_full_run"),
                       full.names = TRUE)
arc_dirs <- arc_dirs[file.info(arc_dirs)$isdir]
hits <- lapply(arc_dirs, function(d) {
  dt <- file.path(d, "hmm", "hmmsearch_domtbl.out")
  if (!file.exists(dt)) return(NULL)
  lines <- readLines(dt); lines <- lines[!grepl("^#", lines)]
  if (length(lines) == 0) return(NULL)
  data.frame(archive = basename(d),
             target  = sub("\\s.*", "", lines),
             stringsAsFactors = FALSE)
})
hits_df <- do.call(rbind, hits)
hits_df |>
  mutate(protein = sub("_ccnbarc.*|_oleic.*", "", target)) |>
  group_by(archive) |>
  summarise(MADA_hits = paste(sort(unique(protein)), collapse = ", "),
            .groups = "drop") |>
  kable(caption = "Proteins that hit MADA per archive (everything else used the DSSP fallback for α1).")
```

In our 14 GT-bearing rows: nbnrc2a and ZAR1 hit MADA in every
archive they appear in; **Sr35 has zero MADA hits even with HMMER
working correctly** (its α1 sequence falls below the MADA HMM's
default detection threshold). RPS2's α1 also did not hit MADA. So
the borderline behaviour of Sr35 in R3 — `SD_THETA_ROT` and
`S_PROTO` close between hexa and penta — is now interpretable as a
real signal, not a tooling artefact: Sr35 *intrinsically* has an α1
that doesn't conform to the MADA pattern, and that's also where some
of its α1-derived features are computed via DSSP rather than HMM.

## Conclusions

### Where we are

After 003 and 004 we have a single artefact that pairs:

- a **curated literature ground truth** of plant NLR oligomeric
  states (`sni_ml/data/known_oligomer_states.csv`, 8 tier-1–3
  cryo-EM entries plus the in-house 18-protein NRC benchmark, all
  graded under the v0.1 rubric locked 2026-04-23), with
- the **Contreras *et al.* 2024 AF3 multimer models** for those NRCs
  computed under controlled forced-stoichiometry conditions (89
  model_0 structures across 10 archives, including the gold
  4/5/6/7/8-mer sweep of NbNRC2a at 10 seeds per stoichiometry),

joined into a **single 44-row feature matrix** at
`sni_ml/results/004_full_run/feature_matrix.csv` carrying all 11
canonical SNI features, ground-truth labels where they exist, and
provenance flags (`archive`, `is_triple_deposit`, `call_correct`).

The 14 GT-bearing rows in R3 and R4 are summarised by eye only —
this notebook does not run any statistical test, plot, or formal
separation analysis. Reading the tables, the values for ZAR1
penta-vs-hexa appear to differ on several features in the direction
we would expect; the NbNRC2a sweep shows the tetramer collapsing
heavily relative to the other stoichiometries; the Sr35 hexa-vs-penta
split looks small. Whether any of this is robust enough to support a
6mer-FP signal is a question for the EDA in **005**, in the same
shape as 001 — distributions, pairwise feature-space plots, and
some explicit measure of separation across the GT-bearing and
GT-less rows. We deliberately do not draw conclusions about feature
discrimination here.

What 004 does establish is the **input for that EDA**: a
GT-anchored, provenance-tagged, replicate-aware feature matrix
covering the 89 Contreras model_0 structures, with the toolchain and
joins documented end-to-end. This is the launch point for 005.

### Decisions carried forward

These are the concrete things this notebook fixes for the rest of
the project. Each is actionable in 005.

1. **Feature matrix is canonical.**
   `results/004_full_run/feature_matrix.csv` is the single source of
   truth for downstream analysis. Re-running this notebook with
   different staging will overwrite it deterministically.
2. **Triple-deposit dedup rule.** Drop
   `archive ∈ {AF_benchmark, Benchmarks}` rows where
   `protein == "nbnrc2a"` and `predicted_oligomer == 6`. Keep the
   `NRC2_hexamers` row.
3. **Sr35 / RPS2 α1 strategy.** Two GT-bearing proteins
   intrinsically miss the MADA HMM (confirmed against
   `hmm/hmmsearch_domtbl.out`, not a tooling artefact); their α1
   features come from the DSSP fallback. In 005 either (a) accept
   the DSSP-fallback values and document the difference, or (b) tune
   the MADA HMM threshold (`-E`/`--incE`) so non-canonical α1s like
   Sr35's hit. Option (a) is the simpler and more defensible default.
4. **`D_MHD_P_mean` decision.** Three options to pick between in
   005, recorded here so the choice is explicit:
   (a) extend `--mhd_csv` and `--ploop_csv` annotations to non-NRC
   proteins via an HMM-driven motif scan;
   (b) accept 10/11 features for the affected rows and document the
   missingness;
   (c) drop the feature from the QC set entirely. Option (a) is the
   most defensible scientifically but takes the most work.
5. **Per-seed analysis path.** The NRC2_*mers `_sd` columns
   summarise between-seed variance, but the per-seed JSONs at
   `results/004_full_run/NRC2_*/per_structure_json/<id>_seed<N>_results.json`
   carry the full per-seed feature values (n = 50 for NbNRC2a alone)
   and are the right input for any seed-aware modelling in 005.

## File index

| Path (relative to `sni_ml/`) | What |
|:--|:--|
| `data/contreras_af3/run_pipeline_all.sh` | per-archive driver — runs `run_pipeline.py` 10× |
| `data/contreras_af3/build_feature_matrix.py` | concatenates per-archive CSVs, parses ID, joins GT |
| `results/004_full_run/<archive>/resistosome_analysis_summary.{csv,xlsx}` | per-archive aggregate (1 row per protein × oligomer) |
| `results/004_full_run/<archive>/per_structure_json/` | per-seed JSON (10 + 1 aggregated per protein × oligomer) |
| `results/004_full_run/<archive>/run.log` | per-archive pipeline stderr/stdout |
| `results/004_full_run/master.log` | driver loop log (per-archive exit codes) |
| `results/004_full_run/feature_matrix.csv` | merged + GT-joined matrix (44 rows × 60+ cols), HMM-correct |
| `results/004_full_run/feature_matrix.xlsx` | same, Excel form |
| `results/004_full_run/feature_matrix_dsspfallback.csv` | frozen first-run matrix where hmmsearch was off PATH; kept as a diff target for the *Methods → Incident* note |
| `results/004_full_run/<archive>/hmm/hmmsearch_domtbl.out` | per-archive MADA HMM hits (which proteins got HMM α1 vs DSSP fallback) |

External resources: the Contreras Zenodo deposit
`10.5281/zenodo.11546022`; the resistosome pipeline lives at
`SolNRCH_foldome/resistosome_pipeline/`; the curated literature
ground-truth table is `sni_ml/data/known_oligomer_states.csv`.
