---
title: "005 — EDA and baseline classifier on the 004 GT-bearing feature matrix"
author: "Dan MacLean"
date: "2026-04-24"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  message: false
  warning: false
---

## Goal

Take the feature matrix produced in notebook 004
(`results/004_full_run/feature_matrix.csv`, 44 rows × ~60 cols), apply
the dedup rule recorded there, restrict to rows where the activated
oligomeric state is known from the literature, and characterise the
canonical 11 SNI features against the binary class

- **TP**: AF3 `predicted_oligomer == gt_oligomer`
- **FP**: AF3 `predicted_oligomer != gt_oligomer`

before fitting any classifier — sample dimensions, class balance,
per-feature missingness, univariate class-separation, within-class
distribution shapes, and inter-feature redundancy. After the EDA, run
the same mlr3 LOOCV + permutation-null pipeline as notebook 002 on
the GT-bearing subset, with α1-feature handling held to a single
strategy because every GT row has the α1 features populated.

## Hypotheses

1. The GT subset after dedup will have a small number of rows (≈12)
   and an imbalanced class ratio (more FP than TP), since the 004
   matrix carries the full NbNRC2a stoichiometry sweep (4 forced-wrong
   stoichiometries vs 1 correct hexamer) plus pentamer/hexamer pairs
   for Sr35, ZAR1, RPS2 (1 correct vs 1 forced for Sr35 and ZAR1, and
   2 forced for RPS2 whose GT is octamer).
2. `D_MHD_P_mean` will be NA across every GT row — and indeed every
   row in the matrix — because the `--mhd_csv` annotation supplied to
   the pipeline only lists NbNRC2a's MHD coordinates, and the
   pipeline emits NA for proteins with no annotation rather than
   imputing. This is the coverage gap flagged in 004 R6 / decision
   list item 4.
3. Of the remaining 10 canonical features, at least
   `S_PROTO`, `D_APEX`, `SD_THETA_ROT` and `BSA_INT_PROTO` will show
   visible TP/FP gaps in the row-wise observations from 004 R3 and
   R4, but at n = 12 with 3 TP / 9 FP no univariate test will reach
   conventional significance after multiple-comparison adjustment.
4. The mlr3 LOOCV pipeline's `featureless` baseline will sit at
   9/12 ≈ 0.75 accuracy by predicting the majority (FP) class on
   every fold. `log_reg` will not converge stably with 10 features and
   12 rows; `ranger` will fit but will likely not exceed the
   permutation null distribution, because the null is constrained to
   the same 3:9 marginal and most label shuffles will themselves
   reach 0.75 accuracy by majority-class prediction on most folds.

## Rationale

n = 12 is small and the class ratio is not balanced, so the EDA is
qualitative for almost every quantity. Despite that, the same set of
EDA passes done in 001 (matrix inventory → missingness →
univariate AUC → distributions → redundancy) is the right lens here
because

- (a) the 004 matrix introduces a different class question (TP vs FP
   AF3 oligomer call) on the same canonical 11 SNI features that 001
   ran on helper vs sensor, so the EDA pattern transfers directly;
- (b) writing the EDA out makes explicit how thin the GT panel is,
   which is the binding constraint on every claim in this and any
   follow-on notebook;
- (c) the baseline classifier section mirrors 002 so that the LOOCV
   accuracy / AUC numbers and the permutation null are reported on
   the same scale the helper/sensor benchmark was reported on,
   making the two readable side-by-side in the writeup.

The classifier section is included for parity with 002 rather than as
a load-bearing result. The conclusions about feature behaviour and
the decisions for 006 are drawn from the EDA, not from the
classifier.

------------------------------------------------------------------------

## Setup

```{r setup}
suppressPackageStartupMessages({
  library(here)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(mlr3)
  library(mlr3learners)
  library(mlr3pipelines)
  library(data.table)
})
lgr::get_logger("mlr3")$set_threshold("warn")

knitr::opts_chunk$set(fig.path = "../results/figures/005_")

fm <- read.csv(here("results", "004_full_run", "feature_matrix.csv"),
               stringsAsFactors = FALSE, na.strings = c("", "NA"))

# pandas wrote booleans as "True"/"False" strings — coerce back
fm$is_triple_deposit <- fm$is_triple_deposit == "True"
fm$call_correct      <- ifelse(fm$call_correct == "True", TRUE,
                        ifelse(fm$call_correct == "False", FALSE, NA))

# 004 dedup rule: keep the NRC2_hexamers nbnrc2a hexamer (n=10 seeds);
# drop the AF_benchmark and Benchmarks duplicates of the same model.
fm_dedup <- fm |>
  filter(!(archive %in% c("AF_benchmark", "Benchmarks") &
           protein == "nbnrc2a" & predicted_oligomer == 6))

gt <- fm_dedup |>
  filter(!is.na(gt_oligomer)) |>
  mutate(class = factor(ifelse(call_correct, "TP", "FP"),
                        levels = c("FP", "TP")),
         row_id = paste(protein, archive, predicted_oligomer, sep = "/"))

CANONICAL_11 <- c("ipTM_LCB", "Sum_CONTACTS", "BSA_INT_PROTO",
                  "SD_THETA_ROT", "S_PROTO", "D_APEX",
                  "THETA_APEX_mean", "L_APEX_mean", "H_ABS_mean",
                  "MU_H_mean", "D_MHD_P_mean")
```

## 1. Matrix inventory

```{r inventory}
dim_tbl <- data.frame(
  matrix = c("004 raw", "004 after triple-deposit dedup",
             "GT-bearing subset (after dedup)"),
  rows   = c(nrow(fm), nrow(fm_dedup), nrow(gt)),
  cols   = c(ncol(fm), ncol(fm_dedup), ncol(gt))
)
knitr::kable(dim_tbl)

table(class = gt$class)

gt |>
  select(row_id, protein, archive, predicted_oligomer, gt_oligomer, class,
         N_replicates) |>
  arrange(class, protein, predicted_oligomer) |>
  knitr::kable()
```

12 rows after dedup, 3 TP and 9 FP, as anticipated. The TP cell is
made of one true hexamer (NbNRC2a from `NRC2_hexamers`, the 10-seed
sweep) and two true pentamers (Sr35 and ZAR1 from `Benchmarks`). The
FP cell is the four forced-wrong NbNRC2a stoichiometries
(tetra/penta/hepta/octa), the NbNRC2a forced-pentamer from
`Benchmarks`, the Sr35 forced-hexamer, the ZAR1 forced-hexamer, and
both RPS2 deposits (whose true state is octamer, so neither the penta
nor the hexa deposit is correct).

## 2. Missingness patterns

### 2.1 Per-feature NA counts by class (canonical 11)

```{r na-counts-11}
na_by_class <- function(df, cols) {
  do.call(rbind, lapply(cols, function(col) {
    data.frame(
      feature  = col,
      n_FP_NA  = sum(is.na(df[df$class == "FP", col])),
      n_TP_NA  = sum(is.na(df[df$class == "TP", col])),
      n_FP     = sum(df$class == "FP"),
      n_TP     = sum(df$class == "TP"),
      row.names = NULL
    )
  }))
}

na_tbl_11 <- na_by_class(gt, CANONICAL_11)
na_tbl_11 <- na_tbl_11[order(-(na_tbl_11$n_FP_NA + na_tbl_11$n_TP_NA)), ]
knitr::kable(na_tbl_11, row.names = FALSE)
```

`D_MHD_P_mean` is NA in every GT row. Looking at the full 44-row
matrix it is NA in every row, full stop, not just in the GT subset:
the `--mhd_csv` annotation supplied to the 004 pipeline only carries
NbNRC2a coordinates and the pipeline emits NA where the lookup
misses. Drop the feature from any per-feature analysis below — the
"canonical 11" reduces to a "canonical 10" for everything that
follows.

The α1 features (`THETA_APEX_mean`, `L_APEX_mean`, `H_ABS_mean`,
`MU_H_mean`) are populated for every GT row, including Sr35 and
RPS2 (whose α1 values come from the DSSP fallback per 004 decision
3a). No row has any other NA in the canonical set.

### 2.2 Missingness grid (canonical 11, GT subset)

```{r na-grid, fig.width=8, fig.height=5}
na_long <- pivot_longer(
  cbind(gt[, c("row_id", "class")], is.na(gt[, CANONICAL_11])),
  cols = all_of(CANONICAL_11),
  names_to = "feature", values_to = "is_na"
)
na_long$feature <- factor(na_long$feature, levels = CANONICAL_11)
na_long$row_id  <- factor(na_long$row_id,
                          levels = gt$row_id[order(gt$class, gt$protein,
                                                   gt$predicted_oligomer)])

ggplot(na_long, aes(x = feature, y = row_id, fill = is_na)) +
  geom_tile(colour = "grey80") +
  scale_fill_manual(values = c(`FALSE` = "white", `TRUE` = "black")) +
  facet_grid(class ~ ., scales = "free_y", space = "free_y") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = NULL, fill = "NA")
```

`D_MHD_P_mean` shows as a black column across both classes; every
other feature is white throughout. From here on the active feature
set is the canonical 10:

```{r canonical-10}
CANONICAL_10 <- setdiff(CANONICAL_11, "D_MHD_P_mean")
CANONICAL_10
```

## 3. Univariate class-separation (rank AUC)

Same Mann–Whitney rank-AUC ranker as notebook 001:
`U / (n_TP * n_FP)`, NAs dropped, positive class = TP. Raw AUC is
directional (high → feature value tends to be larger in TP than FP);
`auc_abs = max(auc, 1 - auc)` is the direction-free separation.

```{r auc-fn}
univariate_auc <- function(x, y, positive = "TP") {
  mask  <- !is.na(x)
  x     <- x[mask]; y <- y[mask]
  n_pos <- sum(y == positive); n_neg <- sum(y != positive)
  if (n_pos < 1 || n_neg < 1) {
    return(data.frame(auc = NA_real_, auc_abs = NA_real_,
                      n_used = length(x), n_pos = n_pos, n_neg = n_neg,
                      wilcox_p = NA_real_))
  }
  r <- rank(c(x[y == positive], x[y != positive]))
  U <- sum(r[seq_len(n_pos)]) - n_pos * (n_pos + 1) / 2
  auc <- U / (n_pos * n_neg)
  p   <- tryCatch(
    suppressWarnings(wilcox.test(x ~ y)$p.value),
    error = function(e) NA_real_
  )
  data.frame(auc = auc, auc_abs = max(auc, 1 - auc),
             n_used = length(x), n_pos = n_pos, n_neg = n_neg,
             wilcox_p = p)
}

auc_table <- function(df, cols) {
  do.call(rbind, lapply(cols, function(col) {
    res <- univariate_auc(df[[col]], df$class)
    cbind(feature = col, res)
  }))
}

auc_10 <- auc_table(gt, CANONICAL_10)
auc_10 <- auc_10[order(-auc_10$auc_abs), ]
knitr::kable(auc_10, row.names = FALSE, digits = 3)
```

Two things to take from this table.

First, **the per-feature separation is much weaker here than in
notebook 001's helper-vs-sensor task on the same canonical features.**
In 001 six of the canonical features (`ipTM_LCB`, `Sum_CONTACTS`,
`BSA_INT_PROTO`, `SD_THETA_ROT`, `S_PROTO`, `D_APEX`) ranked helpers
above sensors *perfectly* — AUC = 1.0. Here only `SD_THETA_ROT`
gets close to that (auc_abs ≈ 0.96), and the rest sit between 0.52
and 0.76. Telling a real-stoichiometry call from a forced-wrong one
is a tighter ask of these features than telling a helper protein
from a sensor protein, and it shows.

Second, **most of the rows that the AUC ranks over are NbNRC2a.**
Five of the 12 GT rows are the NbNRC2a stoichiometry sweep
(tetra/penta/hexa/hepta/octa, 1 TP and 4 FP); a sixth is NbNRC2a
again from `Benchmarks` (FP). So nine of the 12 rows are different
ring sizes of one protein, and only three rows are different
proteins (Sr35, ZAR1, RPS2). The ranking is dominated by
within-protein stoichiometry contrasts — really a kind of
pseudo-replicate — rather than by independent across-protein tests.
A feature that happens to track ring size for NbNRC2a will look
like a strong TP/FP discriminator in this table even if it has
nothing to say about other proteins.

The α1 features sit lower in the ranking because their values
barely move across the NbNRC2a sweep (the α1 helix doesn't really
shift when the ring size changes), so they can't separate
within-protein TP from within-protein FP for the protein that
contributes most of the rows.

## 4. Feature distributions by class

```{r dist-10, fig.width=9, fig.height=8}
long_10 <- pivot_longer(
  gt, cols = all_of(CANONICAL_10),
  names_to = "feature", values_to = "value"
)
long_10$feature <- factor(long_10$feature, levels = CANONICAL_10)

ggplot(long_10, aes(x = class, y = value, colour = class)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.85, size = 2, na.rm = TRUE) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.4, na.rm = TRUE) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_colour_manual(values = c(FP = "#e76f51", TP = "#264653")) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none") +
  labs(x = NULL, y = NULL)
```

With 3 TP points the boxplot whiskers carry essentially no
information — each plot is really 3 TP dots vs 9 FP dots. Read the
panels as scatter, not as box statistics. The features at the top of
the AUC ranking show TP and FP point-clouds that don't overlap; the
features at the bottom show TP dots inside the FP cloud. Keep in
mind when reading the panels that six of the nine FP dots and one
of the three TP dots are the same protein (NbNRC2a) at different
ring sizes — most of the spread within a class is within-protein,
not across proteins.

## 5. Inter-feature redundancy

Spearman, same as 001. The pairs found redundant on the helper/sensor
benchmark in 001 are expected to remain redundant here (the
correlations are between features, not between features and labels).

```{r cor-10, fig.width=7, fig.height=6}
cor_10 <- suppressWarnings(
  cor(gt[, CANONICAL_10], use = "pairwise.complete.obs", method = "spearman")
)

cor_long_10 <- as.data.frame(as.table(cor_10))
names(cor_long_10) <- c("x", "y", "rho")

ggplot(cor_long_10, aes(x = x, y = y, fill = rho)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 2.8) +
  scale_fill_gradient2(low = "#1C39BB", mid = "white", high = "#ED2939",
                       midpoint = 0, limits = c(-1, 1)) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = NULL, fill = expression(rho))
```

```{r cor-pairs-10}
upper_pairs <- function(M, threshold = 0.9) {
  idx <- which(abs(M) >= threshold & upper.tri(M), arr.ind = TRUE)
  if (nrow(idx) == 0) return(data.frame(a = character(), b = character(),
                                        rho = numeric()))
  data.frame(
    a   = rownames(M)[idx[, 1]],
    b   = colnames(M)[idx[, 2]],
    rho = round(M[idx], 3),
    row.names = NULL
  )[order(-abs(M[idx])), ]
}
knitr::kable(upper_pairs(cor_10, 0.9), row.names = FALSE,
             caption = "High-correlation pairs (|rho| >= 0.9)")
```

Two points to keep in mind when reading this matrix: the NbNRC2a
stoichiometry sweep contributes 5 of the 12 rows, all from one
protein, so any pair of features that varies smoothly with chain
count will look strongly correlated here just from that stoichiometry
arc. And with n = 12 the correlation estimates have wide CIs; only
near-perfect correlations should be acted on.

## 6. Baseline classifier (mlr3 LOOCV)

Same scaffolding as notebook 002, restricted to the canonical 10
because `D_MHD_P_mean` is NA throughout. One feature subset
(`canonical_10`) and three learners — `featureless` (no-info
baseline), `log_reg` (linear), `ranger` (random forest, 200 trees).
LOOCV over the 12-row GT subset, scoring accuracy as the per-fold
mean and AUC on the pooled prediction.

```{r design}
make_task <- function(data, features, target = "class") {
  d <- as.data.frame(data[, c(target, features), drop = FALSE])
  TaskClassif$new(id = "sni_qc", backend = d, target = target,
                  positive = "TP")
}

build_learner <- function(lrn_id) {
  base <- lrn(lrn_id, predict_type = "prob")
  if (lrn_id == "classif.ranger") base$param_set$values$num.trees <- 200
  base
}

run_loocv <- function(data, features, lrn_id) {
  tsk <- make_task(data, features)
  lrn_obj <- build_learner(lrn_id)
  rr <- suppressWarnings(resample(tsk, lrn_obj, rsmp("loo")))
  acc_mean <- rr$aggregate(msr("classif.acc"))
  auc_pool <- tryCatch(rr$prediction()$score(msr("classif.auc")),
                       error = function(e) NA_real_)
  list(rr = rr,
       agg = c(classif.acc = unname(acc_mean),
               classif.auc = unname(auc_pool)))
}
```

```{r benchmark}
LEARNERS <- c("classif.featureless", "classif.log_reg", "classif.ranger")

set.seed(1)
bench_results <- do.call(rbind, lapply(LEARNERS, function(lrn_id) {
  res <- tryCatch(
    run_loocv(gt, CANONICAL_10, lrn_id),
    error = function(e) list(agg = c(classif.acc = NA_real_,
                                     classif.auc = NA_real_))
  )
  data.frame(
    cell    = "canonical_10",
    learner = sub("^classif\\.", "", lrn_id),
    acc     = unname(res$agg["classif.acc"]),
    auc     = unname(res$agg["classif.auc"])
  )
}))

knitr::kable(bench_results, digits = 3)
```

`featureless` hits 0.75 accuracy without using any feature values at
all — it just predicts "FP" on every row, and because 9 of the 12
rows are FP, it gets 9 right. This is the baseline to beat.
0.75 on this data means "I learned nothing from the features"; it is
not a 75% real classifier.

`log_reg` and `ranger` are the actual attempts at a classifier using
the 10 SNI features. What we want to see is them scoring clearly
above 0.75. The question of whether they do is settled against the
permutation null in section 7, not by reading this table directly
— because at n = 12 with 9 FP, accuracies near 0.75 happen very
easily under random labels too.

```{r held-out, fig.width=8, fig.height=5}
held_out <- function(data, features, lrn_id) {
  tsk <- make_task(data, features)
  lrn_obj <- build_learner(lrn_id)
  rr <- suppressWarnings(resample(tsk, lrn_obj, rsmp("loo")))
  pred_dt <- as.data.table(rr$prediction())
  pred_dt$row_id <- data$row_id[pred_dt$row_ids]
  pred_dt$class  <- data$class[pred_dt$row_ids]
  pred_dt[, .(row_id, class, response, prob_TP = prob.TP)]
}

ho_ranger <- held_out(gt, CANONICAL_10, "classif.ranger")
ho_ranger$learner <- "ranger"
ho_logreg <- tryCatch({
  d <- held_out(gt, CANONICAL_10, "classif.log_reg")
  d$learner <- "log_reg"
  d
}, error = function(e) NULL)

ho_all <- if (!is.null(ho_logreg)) rbind(ho_ranger, ho_logreg) else ho_ranger
ho_all <- ho_all[order(ho_all$learner, ho_all$class, -ho_all$prob_TP)]
knitr::kable(ho_all, digits = 3)

ggplot(ho_all, aes(x = prob_TP, y = reorder(row_id, prob_TP),
                   colour = class, shape = response == class)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.5) +
  facet_wrap(~ learner) +
  scale_colour_manual(values = c(FP = "#e76f51", TP = "#264653")) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 4),
                     labels = c(`TRUE` = "correct", `FALSE` = "wrong"),
                     name = "prediction") +
  labs(x = "LOOCV held-out P(TP)", y = NULL) +
  theme_minimal(base_size = 10)
```

The held-out plot shows where the per-row predictions sit relative to
the 0.5 boundary. With 12 rows it is short enough to read row by row.

## 7. Permutation null distribution

Class labels are permuted across the 12 GT rows and the LOOCV
pipeline is re-run on each shuffle. `n_perms = 200`, ranger learner.
Empirical one-sided p-value is `(1 + sum(null >= obs)) / (n_perms + 1)`,
floor 1/201 ≈ 0.005.

The shuffling preserves the 3:9 marginal, so the null distribution
itself sits not at 0.5 but at whatever accuracy the ranger pipeline
reaches under random labels with this marginal. The null is the
right calibration for the observed score.

```{r permute}
N_PERMS <- 200
PERM_CACHE <- here("results", "005_permutation_results.rds")

run_permutations <- function(data, features, lrn_id, n_perms, seed) {
  set.seed(seed)
  out <- matrix(NA_real_, nrow = n_perms, ncol = 2,
                dimnames = list(NULL, c("acc", "auc")))
  for (i in seq_len(n_perms)) {
    d_perm <- data
    d_perm$class <- sample(d_perm$class)
    res <- tryCatch(
      run_loocv(d_perm, features, lrn_id),
      error = function(e) list(agg = c(classif.acc = NA_real_,
                                       classif.auc = NA_real_))
    )
    out[i, ] <- c(res$agg["classif.acc"], res$agg["classif.auc"])
  }
  as.data.frame(out)
}

if (file.exists(PERM_CACHE)) {
  perm_results <- readRDS(PERM_CACHE)
  message("Loaded cached permutation results from ", PERM_CACHE)
} else {
  message("Running ", N_PERMS, " permutations (ranger)...")
  t0 <- Sys.time()
  obs   <- run_loocv(gt, CANONICAL_10, "classif.ranger")$agg
  nulls <- run_permutations(gt, CANONICAL_10, "classif.ranger",
                            N_PERMS, seed = 105)
  perm_results <- list(cell = "canonical_10", obs = obs, nulls = nulls)
  saveRDS(perm_results, PERM_CACHE)
  message("Done in ", round(as.numeric(difftime(Sys.time(), t0, units = "secs"))),
          " s")
}
```

```{r permute-table}
perm_summary <- data.frame(
  cell          = perm_results$cell,
  obs_acc       = unname(perm_results$obs["classif.acc"]),
  obs_auc       = unname(perm_results$obs["classif.auc"]),
  null_acc_mean = mean(perm_results$nulls$acc, na.rm = TRUE),
  null_auc_mean = mean(perm_results$nulls$auc, na.rm = TRUE),
  p_acc = (1 + sum(perm_results$nulls$acc >= perm_results$obs["classif.acc"],
                   na.rm = TRUE)) /
          (sum(!is.na(perm_results$nulls$acc)) + 1),
  p_auc = (1 + sum(perm_results$nulls$auc >= perm_results$obs["classif.auc"],
                   na.rm = TRUE)) /
          (sum(!is.na(perm_results$nulls$auc)) + 1)
)
knitr::kable(perm_summary, digits = 3, row.names = FALSE)
```

```{r permute-plot, fig.width=9, fig.height=4}
nulls_long <- data.frame(
  metric = rep(c("acc", "auc"), each = nrow(perm_results$nulls)),
  value  = c(perm_results$nulls$acc, perm_results$nulls$auc)
)
obs_long <- data.frame(
  metric = c("acc", "auc"),
  value  = unname(c(perm_results$obs["classif.acc"],
                    perm_results$obs["classif.auc"]))
)

ggplot(nulls_long, aes(x = value)) +
  geom_histogram(bins = 25, fill = "grey70", colour = "white") +
  geom_vline(data = obs_long, aes(xintercept = value),
             colour = "#ED2939", linewidth = 0.8) +
  facet_wrap(~ metric, scales = "free_x") +
  labs(x = "Score under permuted labels (red = observed)",
       y = "Permutations") +
  theme_minimal(base_size = 10)
```

The permutation histogram centres on about 0.75 accuracy — the
same as the featureless baseline — because any shuffle keeps the
9 FP vs 3 TP split, and predicting "FP for everything" gets 0.75
no matter which rows are labelled TP. The observed ranger score
(red line) sits inside the shuffle pile, not to the right of it.
The p-values in the table make that formal: p_acc ≈ 0.46 and
p_auc ≈ 0.42, both far from significance. **The ranger classifier
fit on the 10 SNI features does not beat random labelling on this
data.** That is the section-7 take-away in one sentence.

## 8. Conclusion

The short version is that **we cannot build a useful TP-vs-FP
classifier from this data**, and three separate things combine to
make that the case.

**1. The per-feature signal is much weaker than in the
helper/sensor benchmark.** In notebook 001, six of the canonical
SNI features separated helpers from sensors *perfectly* (AUC = 1.0).
In 005, on the same canonical features but asking the harder
question "is AF3's stoichiometry call correct," only `SD_THETA_ROT`
gets close to that kind of separation (≈ 0.96); everything else
sits between ~0.55 and ~0.76. So even before we get to sample size,
the features themselves are carrying less class information here
than they did in 001.

**2. The ground-truth panel is very small and dominated by one
protein.** After the dedup rule from 004 we have 12 rows with known
answers, but 9 of them are NbNRC2a (the 5-row stoichiometry sweep
plus the forced-pentamer from `Benchmarks`, plus the "correct"
hexamer) and only 3 rows are other proteins (Sr35 pentamer, ZAR1
pentamer, and two RPS2 deposits whose true state is octameric).
That means most of what looks like TP-vs-FP separation in the AUC
table is really "different ring sizes of the same protein" — a
kind of pseudo-replicate — rather than an independent
across-protein test. The narrower across-protein cut, "given that
AF3 predicted six chains, is that right?", is a n = 4 subset
(1 TP, 3 FP) — too small to get any univariate AUC out of at all.

**3. The classes are unbalanced and the classifier doesn't beat
random labels.** The 3 TP / 9 FP split means a classifier that
ignores the features entirely and always predicts "FP" already
scores 0.75 accuracy just on the class imbalance. Random labellings
of this data also reach 0.75 most of the time, for the same
reason. The ranger classifier fit on the 10 SNI features lands at
the same 0.75 accuracy (AUC 0.33) and has permutation p-values
around 0.46 for accuracy and 0.42 for AUC — the features add
nothing the null doesn't already give.

Put together: **low n, low per-feature signal on this particular
question, and an unbalanced class split where most of the FP side
is pseudo-replicated within one protein.** None of these is the
kind of thing more modelling effort fixes on its own. What would
move the needle is more experimentally-determined NLR oligomeric
states from independent proteins — widening the panel past the
four we currently have — and that is a data-collection problem,
not a classifier problem.

## 9. Decisions / next moves for 006

- **Class definition stays as TP vs FP `call_correct`** for any
  follow-on classifier work in 006. The narrower
  `predicted_oligomer == 6` subset (the 6mer-FP question proper) is
  n = 4 in this matrix and not large enough to fit anything against.
- **Drop `D_MHD_P_mean` from the active feature set** until and
  unless the `--mhd_csv` annotation is extended to non-NRC proteins
  (decision item 4a from 004). 006 should run on the canonical 10.
- **Carry the per-seed JSONs into 006** for the NRC2 archives
  (`results/004_full_run/NRC2_*/per_structure_json/<id>_seed<N>_results.json`):
  the 50 per-seed NbNRC2a rows are the right input for
  any seed-aware modelling, and they raise the effective row count
  for the only protein with a full stoichiometry sweep.
- **Keep `Benchmarks` Sr35 and ZAR1 pentamer rows as TP, hexamer
  rows as FP** — the across-protein TP/FP signal that 006 is
  supposed to test rests on these.

## 10. Session

```{r session}
sessionInfo()
```
