---
title: "007 — Power analysis: how many more GT proteins would we need?"
author: "Dan MacLean"
date: "2026-04-24"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  message: false
  warning: false
---

## Summary

In notebook 004 we joined Contreras *et al.*'s AlphaFold-3 oligomer
predictions to a small set of plant NLRs whose true activated
oligomeric state is known from experiment, giving 12 rows of
"AF3 prediction + correct answer". In 005 we asked the obvious
follow-on: do the canonical SNI structural features tell apart the
rows where AF3 got the oligomer right (TP) from the rows where it
got it wrong (FP)? They didn't — a standard ranger classifier did
no better than always guessing the majority class. In 006 we
checked whether 005's apparent signal was at least transferring
between proteins by training on three proteins and testing on the
fourth; it wasn't, and the classifier collapsed to a baseline majority-class
prediction on every fold.

Our conclusion is that the classification didn't work when we have just a few 'real'
proteins to use.  How much more data would it take before it could?


## Goal

In this document we'll examine features - different SNI values.

- **§1 — How big does each feature's signal look in the current
  data, and how sure are we?** The signal size is measured the same
  way as in notebook 005 (rank-AUC), and the uncertainty is shown
  as a bootstrap 95% CI so we don't pretend the point estimates
  are exact.
- **§2 — How many rows would a standard test need to detect a
  signal that big?** A textbook formula (Noether 1987) converts
  effect size + class ratio into "minimum sample size for 80%
  power." We compute it for every feature.
- **§3 — The same calculation as a curve.** Power vs sample size
  for the top three features, with the current `n = 12` and the
  80% power target both drawn on the plot, so the gap between
  where we are and where we'd want to be is visible at a glance.
- **§4 — From rows to proteins.** A sample-size in rows is one
  unit; the more useful unit for future work is "how many
  new ground-truth proteins do we need to find?". This section
  does that conversion.
- **§5 — Plain-language conclusion.**
- **§6 — What we'll do next.**

### Some terms

- **AUC** (here, `auc_abs`) — Area Under the ROC Curve, a number
  between 0.5 and 1.0 that says how well a feature's values
  separate two classes when used as a 1-D ranker. **0.5** means
  "no better than a coin flip — TP and FP rows are interleaved";
  **1.0** means "every TP value is above every FP value (or vice
  versa)". `auc_abs = max(AUC, 1 − AUC)` so the number is
  direction-free.
- **Power** — the probability that, if the feature really does
  separate the classes by the amount we observed, a standard test
  on a fresh sample of size `n` would correctly detect it. The
  conventional target is 80%: "in 4 out of 5 datasets of this
  size, the test would catch the signal."
- **α** — the false-positive rate of the test. α = 0.05 is the
  conventional threshold: "we'll only call a result significant if
  it would happen by chance in fewer than 1 in 20 datasets when
  there is actually no real signal."
- **Bootstrap 95% CI** — because we only have 12 rows, the AUC
  point estimate is itself uncertain. The bootstrap re-samples the
  12 rows with replacement many times, recomputes the AUC each
  time, and reports the 2.5%–97.5% range as a 95% confidence
  interval. A wide CI means "the true AUC could be quite different
  from the point estimate; the data we have can't pin it down."

## Hypotheses

Stated upfront so the reader can check whether what comes out of
the analysis matches what we expected going in.

1. **The bootstrap CIs on per-feature AUC will be wide,** wide
   enough that for most features the CI overlaps AUC = 0.5
   (no real signal). This is the concrete version of "12 rows,
   imbalanced 3:9, can't precisely pin down an effect size."
2. **`SD_THETA_ROT` will look strong on the point estimate** (≈ 0.96
   in 005), so the sample-size formula will say we already have
   enough rows. But its CI will be wide, and we already know from
   notebook 006 that this feature's signal doesn't transfer across
   proteins — so the headline number for this feature is a
   misleading one to act on.
3. **The middle-ranked features** (`ipTM_LCB`, `S_PROTO`,
   `BSA_INT_PROTO`, `D_APEX`, AUC ≈ 0.67–0.76) will need on the
   order of **tens of independent proteins** for 80% power.
4. **The weakest features** (AUC ≈ 0.55) will need **hundreds to
   thousands** of proteins, which isn't a realistic target for any
   plant-NLR curation effort.


We deliberately don't run a multivariate classifier power analysis
here. At `n = 12` the per-class feature distributions are
themselves too poorly estimated for a simulation to be meaningful;
any classifier-power numbers it produced would just be reading
back the assumptions we'd had to invent. The univariate Wilcoxon
power analysis is the defensible lower bar: if even a
single-feature test needs 50+ proteins, a multivariate classifier
on the same features won't beat it on fewer.

------------------------------------------------------------------------

## Setup

Same load, same dedup rule, same canonical 10 features as 005 and 006.

```{r setup}
suppressPackageStartupMessages({
  library(here)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(WMWssp)
})

knitr::opts_chunk$set(fig.path = "../results/figures/007_")

fm <- read.csv(here("results", "004_full_run", "feature_matrix.csv"),
               stringsAsFactors = FALSE, na.strings = c("", "NA"))
fm$is_triple_deposit <- fm$is_triple_deposit == "True"
fm$call_correct      <- ifelse(fm$call_correct == "True", TRUE,
                        ifelse(fm$call_correct == "False", FALSE, NA))

fm_dedup <- fm |>
  filter(!(archive %in% c("AF_benchmark", "Benchmarks") &
           protein == "nbnrc2a" & predicted_oligomer == 6))

gt <- fm_dedup |>
  filter(!is.na(gt_oligomer)) |>
  mutate(class = factor(ifelse(call_correct, "TP", "FP"),
                        levels = c("FP", "TP")))

CANONICAL_10 <- c("ipTM_LCB", "Sum_CONTACTS", "BSA_INT_PROTO",
                  "SD_THETA_ROT", "S_PROTO", "D_APEX",
                  "THETA_APEX_mean", "L_APEX_mean", "H_ABS_mean",
                  "MU_H_mean")
```

## 1. How big is each feature's signal how variable is it?

The signal size is the rank-AUC of the feature (SNI value) treated as a
single-feature classifier of TP vs FP, exactly as in notebook 005
§3. The uncertainty around each estimate is the bootstrap 95% CI
described in the jargon section. The plot at the end is the more
important artefact than the table — it's where the reader sees that
the point estimates are not as solid as they look.

```{r auc-boot}
auc_abs <- function(x, y, pos = "TP") {
  m <- !is.na(x); x <- x[m]; y <- y[m]
  np <- sum(y == pos); nn <- sum(y != pos)
  if (np < 1 || nn < 1) return(NA_real_)
  r  <- rank(c(x[y == pos], x[y != pos]))
  U  <- sum(r[seq_len(np)]) - np * (np + 1) / 2
  a  <- U / (np * nn)
  max(a, 1 - a)
}

boot_auc_ci <- function(gt, col, n_boot = 2000, seed = 7) {
  set.seed(seed)
  out <- replicate(n_boot, {
    idx <- sample(nrow(gt), replace = TRUE)
    auc_abs(gt[[col]][idx], gt$class[idx])
  })
  c(lo = unname(quantile(out, 0.025, na.rm = TRUE)),
    hi = unname(quantile(out, 0.975, na.rm = TRUE)))
}

auc_tbl <- do.call(rbind, lapply(CANONICAL_10, function(c) {
  ci <- boot_auc_ci(gt, c)
  data.frame(feature = c,
             auc_abs = auc_abs(gt[[c]], gt$class),
             ci_lo   = ci["lo"], ci_hi = ci["hi"])
}))
auc_tbl <- auc_tbl[order(-auc_tbl$auc_abs), ]
knitr::kable(auc_tbl, digits = 3, row.names = FALSE)
```

```{r auc-ci-plot, fig.width=7, fig.height=4}
ggplot(auc_tbl, aes(x = auc_abs, y = reorder(feature, auc_abs))) +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.25,
                 colour = "grey50") +
  geom_point(size = 3, colour = "#264653") +
  labs(x = "auc_abs (0.5 = no effect, 1.0 = perfect)",
       y = NULL,
       subtitle = "Point estimate + bootstrap 95% CI, n = 12 rows") +
  xlim(0.4, 1.05) +
  theme_minimal(base_size = 10)
```

The dot is the AUC we measured on the 12 rows of SNI data
the horizontal grey bar is the 95% CI from bootstrapping. The
vertical dashed line at 0.5 marks "no effect." A feature whose CI
crosses or touches the 0.5 line is one where, given how little data
we have, the data themselves can't rule out "this feature has no
real signal."

What we see. The point-estimate ranking is the same as in 005's
table — `SD_THETA_ROT` at the top, the α1 features at the bottom.
What's new here is that almost every feature's CI touches or
crosses 0.5. Only `SD_THETA_ROT`'s CI clearly excludes 0.5, and even
its CI runs all the way up to 1.0. The take-away from §1 is the
honest one: **every sample-size number we compute in §2 is
conditional on the point estimate being correct, and the point
estimates are uncertain.**

## 2. How many rows would a test need to detect a signal that size?

This is where the power analysis happens. The Noether (1987)
formula converts an effect size into a sample-size requirement for
the Wilcoxon rank test:

$$
n_{\text{total}} = \frac{(z_{\alpha/2} + z_\beta)^2}{c\,(1-c)\cdot 12\cdot (p - 0.5)^2}
$$

where `p` is the AUC, `c` is the fraction of rows in the positive
class, `z_{α/2} = 1.96` for α = 0.05 (two-sided), and `z_β = 0.84`
for 80% power. We don't roll our own implementation — we use the
`WMWssp` package's `WMWssp_noether()`, which is a direct
implementation of this formula.

The intuition is the bit worth carrying away from the algebra. The
key part is `(p − 0.5)²` in the denominator: **bigger gap from
chance → fewer samples needed; smaller gap → quadratically more
samples**. Halving the AUC gap (e.g. going from AUC 0.80 to AUC
0.65 — same 0.15 distance from 0.5 either side) quadruples the
required sample size. That's the reason a feature with
a weak-but-real signal can need a hundred times more data than a
feature with a strong signal.

The class-ratio piece, `c(1 − c)`, is maximised at `c = 0.5`
(balanced) and gets worse as the classes become unbalanced. That's
why the table below shows two columns: the balanced-ideal case
(1:1) and the unbalanced case we'd actually face if we kept the
current 1 TP : 3 FP ratio.

```{r noether}
ssp_n <- function(auc, c_ratio, alpha = 0.05, power = 0.80) {
  WMWssp_noether(alpha = alpha, power = power,
                 t = c_ratio, p = auc)$N
}

n_tbl <- auc_tbl |>
  rowwise() |>
  mutate(n_total_1to1 = ssp_n(auc_abs, 0.5),
         n_total_1to3 = ssp_n(auc_abs, 0.25)) |>
  ungroup()
knitr::kable(
  n_tbl[, c("feature", "auc_abs", "n_total_1to1", "n_total_1to3")],
  digits = 3, row.names = FALSE,
  col.names = c("feature", "auc_abs",
                "n_total (1:1 balanced)",
                "n_total (1:3 TP:FP)"))
```

The features split cleanly into three bands at the observed point
estimates.

- **Already within reach.** `SD_THETA_ROT` (AUC ≈ 0.96) needs only
  ~14–18 rows; we already have 12. The asterisk is enormous,
  though: as §1 showed, the CI on this AUC is wide, and as 006
  showed, the apparent signal is dominated by the NbNRC2a
  stoichiometry sweep and doesn't generalise across proteins. The
  formula's verdict on this feature is right arithmetically but
  the wrong thing to act on biologically.
- **Tens of proteins away.** `ipTM_LCB`, `S_PROTO`, `BSA_INT_PROTO`,
  `D_APEX` (AUC ≈ 0.67–0.76) need ~40–130 total rows, depending on
  the class ratio. Under the rough rule that each newly-curated
  protein contributes one informative row (see §4), this means
  **on the order of 40–130 more independent GT proteins**.
- **Hundreds to thousands of rows.** `Sum_CONTACTS`, `MU_H_mean`,
  and the three α1 features (AUC ≈ 0.52–0.63) need 200 to 10000+
  rows. These are not realistically reachable by curating more
  plant NLRs; they will not become detectable single-feature
  classifiers on this task.

## 3. The same calculation as a curve

The table in §2 gives a number per feature, but a single number is
hard to put into context. The plot below shows the same calculation
as a curve — power as a function of total sample size — for the
top three features. Two reference lines are drawn so the reading is
direct: a horizontal dashed line at 80% power (the conventional
target) and a vertical dashed line at our current `n = 12`.

`WMWssp_noether()` returns the sample size needed for a target
power; it doesn't return the power achievable at a given sample
size. We get the curve by calling the package the way it wants to
be called: sweep the target power from 0.05 to 0.999, ask the
package what N that would require at each level, and plot the
resulting (N, power) pairs. The package does the arithmetic.

```{r power-curves, fig.width=7, fig.height=4.5}
top3 <- head(auc_tbl$feature, 3)
power_grid <- seq(0.05, 0.999, by = 0.005)

grid <- do.call(rbind, lapply(top3, function(f) {
  a <- auc_tbl$auc_abs[match(f, auc_tbl$feature)]
  data.frame(
    feature = f,
    auc     = a,
    power   = power_grid,
    n_total = sapply(power_grid, function(pw)
      WMWssp_noether(alpha = 0.05, power = pw,
                     t = 0.25, p = a)$N)
  )
}))
grid <- grid[grid$n_total <= 200, ]

ggplot(grid, aes(x = n_total, y = power, colour = feature)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 12,  linetype = "dashed", colour = "grey70") +
  geom_line(linewidth = 0.8) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Total sample size (at TP:FP = 1:3)", y = "Power at α = 0.05",
       subtitle = "Dashed: 80% power target (horizontal), current n = 12 (vertical)") +
  theme_minimal(base_size = 10)
```

The plot shows at what point the number of replicates gives you an 80%
chance of classifying correctly. Find the feature you care about (curve
colour), trace it across the x-axis until it crosses the
horizontal 80% line, then read the x-value beneath the crossing.
That x-value is the total sample size that feature would need for
a Wilcoxon test to detect its signal in 4 out of 5 hypothetical
datasets at the AUC we currently observe.

What we see. At the current `n = 12` (vertical dashed), only
`SD_THETA_ROT` is above the 80% target line.
`ipTM_LCB` crosses 80% at around `n ≈ 50`; `S_PROTO` crosses it at
around `n ≈ 60`. Both crossings assume the AUC point estimate is
the truth, which §1's CIs say we shouldn't believe too tightly.

## 4. From rows to proteins: how many new ones to curate?

The numbers in §2 and §3 are in *rows*. A row count isn't directly
useful for a curation plan — what we actually need to know is
"how many new proteins" — so this section translates.

If we assume independent proteins (not very close orthologues),
the realistic assumption is **each new GT protein
contributes about one informative row** — typically the AF3
prediction of its activated state, classified as TP or FP against
the experimental result. Under that rule of thumb, the row counts
in §2 convert as follows:

```{r row-to-protein}
convert_tbl <- n_tbl |>
  mutate(n_proteins_1to3 = n_total_1to3,
         n_new_proteins  = pmax(n_total_1to3 - 4, 0)) |>
  select(feature, auc_abs, n_total_1to3, n_proteins_1to3, n_new_proteins)
knitr::kable(convert_tbl, digits = 3, row.names = FALSE,
             col.names = c("feature", "auc_abs",
                           "n total rows (1:3)",
                           "≈ n proteins",
                           "≈ n new proteins on top of 4"))
```

The right-hand column is the practically useful one: **how many
plant NLRs with experimentally-resolved activated oligomeric state
would we need to add to the four we already have to make each
feature individually detectable** at 80% power.

## 5. Summary

** The data we have right now can't precisely measure the
effect sizes themselves.** Almost every per-feature AUC's 95% CI
touches or crosses 0.5 (no effect) — including some of the higher
point estimates. So every sample-size number in §2 and §3 is
conditional on the point-estimate AUC being right, and at `n = 12`
the point estimates are not pinned down. Treat the numbers as
"order of magnitude" guides, not as exact targets.

**`SD_THETA_ROT`'s "we already have enough" verdict is real
arithmetically but misleading biologically.** The formula is
correct on its own terms, but most of the apparent signal comes
from contrasting NbNRC2a's correct hexamer against four
forced-wrong stoichiometries of the *same protein*. Notebook 006
showed that signal doesn't transfer across proteins. Curating more
of the same — only NbNRC2a — would not change this. What's needed
for `SD_THETA_ROT` is independent proteins, not more rows of the
one we have.

**For the middle band of features (`ipTM_LCB`, `S_PROTO`,
`BSA_INT_PROTO`, `D_APEX`, AUC ≈ 0.67–0.76), the realistic target
is roughly 40–130 independent GT proteins on top of the 4 we
have.** This is a non-trivial but reachable curation programme,
plausibly addressable in the near term.

**The weakest features won't become useful single-feature
classifiers no matter how much more data we collect.** When the
required sample is in the hundreds to thousands of independent
proteins, the answer for a structural-biology curation programme
is "this feature is not really good for this task." Of course, these
numbers are not super strong, but it is suggestive that these features
are not very useful in the SNI overall.



## 7. Session

```{r session}
sessionInfo()
```
